---
title: "Environmental Kernels and Auditable Interaction Hypotheses"
output: rmarkdown::html_vignette
header-includes:
  - |
    <script>
    window.MathJax = { tex: { inlineMath: [['\\(','\\)']], displayMath: [['\\[','\\]']] }, svg: { fontCache: 'global' } };
    </script>
    <script defer src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
vignette: >
  %\VignetteIndexEntry{Environmental Kernels and Auditable Interaction Hypotheses}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# 1. Purpose

Environmental information can improve the design of a multi-environment trial
(MET), but environmental similarity is not genetic covariance. Moreover,
stating that a Gaussian kernel represents interactions does not establish that
a particular combination, such as temperature by root-zone bulk density,
matters to genotype performance.

OptiSparseMET therefore separates: (i) nonlinear similarity within one
environmental modality, (ii) omnibus interactions between whole modalities,
and (iii) dedicated interactions among named variables. This distinction
supports robust design without converting an untested physiological hypothesis
into a fitted genetic relationship.

The objectives of this vignette are to: (i) define the three kernel levels,
(ii) document every required and optional input, (iii) explain strong heredity
and feature residualisation, (iv) show how historical response evidence is
used, and (v) specify the no-history fallback.

# 2. Three environmental representations

## 2.1 Modality main kernels

Weather, soil, management, and geography are retained as separate evidence
blocks. For a non-geographic modality \(m\), the default kernel is:

\[
K_m(e,e') =
\frac{1}{|B|}
\sum_{b\in B}
\exp\left[
-\frac{\lVert z_{e,m}-z_{e',m}\rVert^2}{2b^2}
\right],
\]

where \(z_{e,m}\) is the vector of standardised covariates for environment
\(e\), and \(B\) contains bandwidths derived from the median positive
pairwise distance. The default multipliers are 0.5, 1, and 2.

The Gaussian kernel represents nonlinear functions of all variables within
the modality. Its feature space consequently contains implicit interactions.
However, it does not estimate or report a named coefficient for temperature by
rainfall, temperature by humidity, or any other variable pair.

## 2.2 Omnibus modality interactions

When `include_interactions = TRUE`,
`build_environment_kernels()` constructs every available pair among weather,
soil, and management. The default functional analysis of variance
(functional-ANOVA) interaction is:

\[
K_{a\times b} = (H K_a H)\circ(H K_b H),
\]

where \(H=I-\mathbf{1}\mathbf{1}'/n\) and \(\circ\) denotes the Hadamard
product. Centring removes the constant feature-space baseline from each
parent. The resulting tensor-product kernel is positive semidefinite.

An omnibus weather-by-soil kernel represents functions of all retained weather
variables jointly with functions of all retained soil variables. Hence, it
contains temperature-by-bulk-density functions when both variables survived
quality control. Nevertheless, it cannot identify that pair separately from
rainfall by clay content, vapour-pressure deficit by available water, or the
other combinations in the same feature spaces.

## 2.3 Dedicated variable interactions

A dedicated kernel is appropriate when an interaction must be guaranteed,
audited, and assessed separately. Examples include:

- reproductive-stage temperature by root-zone bulk density;
- rainfall by root-zone clay content;
- nitrogen dose by water availability;
- sowing date by thermal time; and
- humidity by temperature for disease-conducive conditions.

These hypotheses should be specified before examining the historical response
results whenever possible. Exhaustive generation of every variable pair is not
recommended because the number of candidate terms grows rapidly and the
environment sample is usually small.

# 3. Required and optional inputs

## 3.1 Quality-controlled covariate blocks

`build_variable_interaction_kernels()` requires a named list of numeric
environment-by-variable matrices. The `covariates` element returned by
`build_environment_kernels()` is the recommended input because identifiers,
missing values, categorical expansion, and variable filtering have already
been handled.

Every block must satisfy the following requirements:

| Requirement | Rule |
|---|---|
| Environment identifiers | Unique, non-missing row names |
| Environment coverage | The same environments in every block |
| Variable identifiers | Unique, non-empty column names |
| Values | Finite numeric values after quality control |
| Categorical factors | One-hot encoded before dedicated kernel construction |

Example weather block:

| environment | T2M_mean_reproductive | PRECTOTCORR_sum_reproductive | VPD_q90 |
|---|---:|---:|---:|
| E1 | 26.8 | 318.4 | 1.72 |
| E2 | 27.4 | 212.7 | 2.06 |
| E3 | 24.1 | 286.3 | 1.48 |

Example soil block:

| environment | bulk_density_root | clay_root | available_water_capacity_root |
|---|---:|---:|---:|
| E1 | 1.28 | 34.2 | 0.164 |
| E2 | 1.41 | 27.6 | 0.132 |
| E3 | 1.19 | 41.5 | 0.188 |

## 3.2 Interaction specification

The recommended specification is a long data frame with four required
columns:

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `interaction` | character | yes | Unique interaction-hypothesis label |
| `parent` | character | yes | Parent feature label within that interaction |
| `modality` | character | yes | Covariate-block name |
| `variable` | character | yes | Exact post-quality-control column name |

Example:

| interaction | parent | modality | variable |
|---|---|---|---|
| temperature_x_density | temperature | weather | T2M_mean_reproductive |
| temperature_x_density | density | soil | bulk_density_root |
| rainfall_x_clay | rainfall | weather | PRECTOTCORR_sum_reproductive |
| rainfall_x_clay | clay | soil | clay_root |
| nitrogen_x_water | nitrogen | management | nitrogen_kg_ha |
| nitrogen_x_water | water | weather | water_balance_reproductive |

Each interaction requires at least two distinct parents. Several rows can share
the same interaction and parent. This defines a multicolumn parent, which is
useful for a categorical management factor represented by several indicator
columns.

The equivalent named-list format is:

```{r interaction-list-format, eval=FALSE}
interaction_hypotheses <- list(
  temperature_x_density = list(
    temperature = list(
      modality = "weather",
      variables = "T2M_mean_reproductive"
    ),
    density = list(
      modality = "soil",
      variables = "bulk_density_root"
    )
  )
)
```

## 3.3 Optional construction controls

| Argument | Default | Purpose |
|---|---|---|
| `bandwidth_multipliers` | `c(0.5, 1, 2)` | Average several Gaussian bandwidths |
| `orthogonalise` | `TRUE` | Residualise the tensor feature against its parents |
| `hierarchy` | `"strong"` | Retain exact parent-variable kernels |
| `max_tensor_columns` | `200` | Prevent unnoticed categorical expansion |
| `min_residual_fraction` | `1e-6` | Reject an interaction almost completely explained by its parents |

`build_environment_kernels()` exposes these controls through one compact
argument, `variable_interaction_control`.

# 4. Construction method

## 4.1 Parent features

For interaction \(h\), let \(X_{h1},\ldots,X_{hq}\) denote the \(q\) parent
feature blocks. Each column is centred and scaled. Non-informative columns are
rejected because they cannot distinguish environments.

## 4.2 Row-wise tensor product

The raw interaction feature matrix contains every row-wise product across the
parent blocks. If two parents contain \(p_1\) and \(p_2\) columns, the raw
two-way tensor contains \(p_1p_2\) columns. Three parents contain
\(p_1p_2p_3\) columns. Therefore, the `max_tensor_columns` guard is a
statistical and computational requirement rather than a convenience.

## 4.3 Parent-effect residualisation

Let \(Q_h\) be the standardised tensor feature matrix and
\(P_h=[\mathbf{1},X_{h1},\ldots,X_{hq}]\) the parent design. The default
residual interaction feature is:

\[
Q_h^\perp =
\left[I-P_h(P_h'P_h)^{-}P_h'\right]Q_h.
\]

This removes the component linearly explained by the intercept and parent
features. Residualisation occurs before the Gaussian kernel is constructed.
Therefore, the final kernel remains positive semidefinite. The audit reports:

\[
r_h =
\frac{\lVert Q_h^\perp\rVert_F^2}
     {\lVert Q_h\rVert_F^2},
\]

the fraction of tensor-feature information retained after residualisation.
An extremely small value indicates aliasing or inadequate environmental
variation.

Residualisation reduces main-effect leakage but does not make a causal claim.
Nor does it guarantee independence from every nonlinear function in the broad
modality kernels.

## 4.4 Dedicated Gaussian kernel

The residual feature matrix is standardised again and converted to a Gaussian
bandwidth ensemble. This provides an exact, separately named environmental
similarity for the stated interaction. The kernel is symmetric, positive
semidefinite, and normalised.

## 4.5 Strong heredity

Strong heredity means that an interaction is never assessed without its exact
parent-variable kernels. This rule prevents an apparent interaction from
substituting for an omitted parent main effect. It is structural: a historical
simplex calibration may assign a very small weight to a parent because kernel
weights are covariance contributions, not regression coefficients.

# 5. End-to-end construction

```{r build-dedicated-interactions, eval=FALSE}
library(OptiSparseMET)

interaction_hypotheses <- data.frame(
  interaction = c(
    "temperature_x_density", "temperature_x_density",
    "rainfall_x_clay", "rainfall_x_clay"
  ),
  parent = c("temperature", "density", "rainfall", "clay"),
  modality = c("weather", "soil", "weather", "soil"),
  variable = c(
    "T2M_mean_reproductive", "bulk_density_root",
    "PRECTOTCORR_sum_reproductive", "clay_root"
  )
)

environment_model <- build_environment_kernels(
  weather = weather_features,
  soil = soil_features,
  management = management_features,
  geography = site_coordinates,
  include_interactions = TRUE,
  variable_interactions = interaction_hypotheses,
  variable_interaction_control = list(
    hierarchy = "strong",
    orthogonalise = TRUE,
    max_tensor_columns = 100L
  )
)

names(environment_model$kernels)
environment_model$variable_interaction_audit
environment_model$variable_ledger
environment_model$kernel_hierarchy
```

The broad and dedicated kernels coexist. For example,
`weather_x_soil` remains the omnibus uncertainty structure, whereas
`variable_x__temperature_x_density` represents the named variable
hypothesis.

# 6. Audit and failure conditions

The interaction audit contains:

| Field | Interpretation |
|---|---|
| `interaction` | User-defined hypothesis label |
| `kernel` | Realised safe kernel name |
| `order` | Number of parent features |
| `scope` | Within-modality or cross-modality |
| `modalities` | Ordered parent modalities |
| `variables` | Exact matched covariate columns |
| `parent_kernels` | Strong-heredity parent mapping |
| `raw_tensor_columns` | Dimension before filtering |
| `informative_tensor_columns` | Informative dimension before residualisation |
| `final_tensor_columns` | Informative residual dimension |
| `residual_information_fraction` | Information retained beyond parent main effects |
| `effective_kernel_rank` | Numerically informative kernel dimensions |
| `bandwidths` | Realised Gaussian bandwidth ensemble |
| `status` | Structural-construction result |

Construction stops when:

1. A modality is absent.
2. A named variable did not survive quality control.
3. A parent contains no informative variation.
4. The raw tensor exceeds `max_tensor_columns`.
5. The tensor is constant.
6. Residualisation retains less than `min_residual_fraction`.
7. Distinct definitions produce colliding safe kernel names.

These failures should not be bypassed automatically. They indicate either an
input mismatch or an interaction that cannot be distinguished in the available
environment sample.

# 7. Known-truth interaction demonstration

A simulation can verify whether the calibration recovers an interaction that
is known to generate environmental genetic covariance. The simulated weights
are part of the data-generating mechanism; they are not user-defined priors for
a real no-history analysis.

```{r known-truth-variable-interaction, eval=FALSE}
variable_only <- build_variable_interaction_kernels(
  covariates = environment_model$covariates,
  interactions = interaction_hypotheses[1:2, ]
)

true_weights <- stats::setNames(
  c(0.20, 0.20, 0.50, 0.10),
  c(names(variable_only$kernels), "identity")
)

known_truth <- simulate_environment_model_benchmark(
  kernels = variable_only$kernels,
  true_weights = true_weights,
  n_genotypes = 300,
  n_years = 12,
  sigma_g2 = 1,
  sigma_e2 = 1,
  reps_per_cell = 2,
  missing_rate = 0.05,
  n_boot = 500,
  seed = 2026
)

known_truth$weight_recovery
known_truth$benchmark$scores
known_truth$benchmark$cross_validation
```

The benchmark should report: (i) recovery of the known parent and interaction
weights, (ii) off-diagonal root-mean-square error and relative Frobenius error,
(iii) leave-year-out and leave-environment-out prediction, and (iv) bootstrap
weight uncertainty. Recovery should improve as the number of genotypes, years,
and informative environment pairs increases. Failure under a realistic sample
size indicates that the interaction is not identifiable enough for central
weighting, even though it can remain a robustness scenario.

# 8. Historical response evidence

## 8.1 Input format

Historical responses are optional. When supplied, they should be adjusted
genetic values, best linear unbiased estimators (BLUEs), or best linear
unbiased predictors (BLUPs), not unadjusted plot observations.

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `genotype` | character | yes | Historical genotype identifier |
| `environment` | character | yes | Environment identifier aligned to the kernels |
| `value` | numeric | yes | Adjusted genetic value for one target trait |
| `year` | integer or character | no | Defines year-specific targets and leave-year-out validation |

Example:

| genotype | environment | value | year |
|---|---|---:|---:|
| G001 | E1 | 0.84 | 2023 |
| G001 | E2 | 0.31 | 2023 |
| G002 | E1 | -0.12 | 2023 |
| G002 | E2 | 0.18 | 2023 |

## 8.2 Component-wise assessment

```{r assess-dedicated-interactions, eval=FALSE}
interaction_evidence <- assess_variable_interactions(
  kernels = environment_model$kernels,
  historical = adjusted_historical_met,
  genotype_col = "genotype",
  environment_col = "environment",
  value_col = "value",
  year_col = "year",
  alpha = 0.05,
  p_adjust = "holm",
  n_boot = 500,
  seed = 2026
)

interaction_evidence$evidence
interaction_evidence$supported
interaction_evidence$structural_uncertainty
```

Every dedicated interaction is compared separately with the same main-effect
model. Leave-year-out validation is preferred when several year-specific
targets exist. Otherwise, leave-environment-out validation is used.

An interaction is supported only when:

1. Its mean blocked-validation root-mean-square error improvement is positive.
2. The one-sided paired sign test is significant after the selected
   multiplicity correction.
3. Its fitted covariance contribution is positive.

Holm correction is the default because it controls the family-wise error rate
without requiring independent tests. `p_adjust = "none"` is defensible only
for one pre-registered hypothesis or an explicitly exploratory analysis.

Bootstrap inclusion is the fraction of pair-bootstrap fits in which the
interaction received a positive covariance contribution. It quantifies
weight-selection stability but does not replace blocked prediction.

## 8.3 Central covariance and uncertainty candidates

```{r calibrate-screened-interactions, eval=FALSE}
environment_covariance <- calibrate_environment_covariance(
  kernels = environment_model$kernels,
  historical = adjusted_historical_met,
  year_col = "year",
  interaction_kernels =
    interaction_evidence$evidence$interaction,
  eligible_interactions = interaction_evidence$supported,
  interaction_policy = "evidence",
  n_boot = 500,
  seed = 2026
)

Sigma_E <- environment_covariance$Sigma_E
environment_covariance$interaction_evidence
names(environment_covariance$candidates)
```

`eligible_interactions` controls entry into the central covariance only.
Screened interactions remain in `candidates` and therefore remain available
for robust sensitivity analysis. The aggregate evidence gate then verifies
that the supported components improve prediction jointly.

# 9. No historical MET responses

```{r no-history-interactions, eval=FALSE}
interaction_evidence <- assess_variable_interactions(
  kernels = environment_model$kernels
)

environment_covariance <- calibrate_environment_covariance(
  kernels = environment_model$kernels
)
```

The no-history branch has four explicit consequences:

1. `Sigma_E` is the identity matrix.
2. Kernel weights are `NULL`.
3. No dedicated interaction is labelled supported or unsupported.
4. Every main, omnibus interaction, and dedicated interaction kernel remains a
   separate unweighted structural scenario.

Designs should be compared by maximin performance across these structures.
Equal weights or user-defined probabilities must not be introduced merely to
obtain one combined kernel.

# 10. What the method can and cannot establish

The method can establish that:

- every requested variable survived quality control;
- a named interaction was represented separately;
- its tensor feature contained information beyond the parent linear features;
- its kernel was positive semidefinite and numerically informative;
- historical blocked prediction improved, when sufficient response data exist;
- the conclusion was or was not stable across bootstrap resamples; and
- the released design remains acceptable when the interaction is treated as a
  structural uncertainty scenario.

The method cannot establish that:

- an implicit Gaussian feature corresponds to one identifiable biological
  mechanism;
- a historical association is causal;
- an unsupported interaction is physically absent;
- a supported interaction will transport to a new target population of
  environments without prospective validation; or
- a small environment network can support exhaustive interaction discovery.

# 11. Recommended breeding workflow

Use the following sequence:

1. Define the target population of environments and the target trait.
2. Quality-control weather, soil, management, and geographic inputs.
3. Construct separate modality main kernels.
4. Retain omnibus modality interactions as broad structural scenarios.
5. Pre-specify a short list of physiological or management interactions.
6. Construct dedicated residualised kernels with strong heredity.
7. Review the variable ledger, residual information, effective rank, and
   tensor dimension.
8. Assess each dedicated interaction with blocked historical response evidence
   when available.
9. Control multiplicity and evaluate bootstrap stability.
10. Admit supported interactions to the aggregate central-covariance gate.
11. Retain all broad, weakly supported, and rejected interactions as separate
    maximin scenarios.
12. Report both the central recommendation and the structural sensitivity
    results before releasing the field design.

Hence, environmental interactions inform the design without being confused
with established genetic covariance or causal environmental mechanisms. This
is the appropriate distinction for an auditable breeding decision.
