---
title: "Benchmarking and Validation of OptiSparseMET Designs"
output: rmarkdown::html_vignette
header-includes:
  - |
    <script>
    window.MathJax = { tex: { inlineMath: [['\\(','\\)']], displayMath: [['\\[','\\]']] }, svg: { fontCache: 'global' } };
    </script>
    <script defer src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
vignette: >
  %\VignetteIndexEntry{Benchmarking and Validation of OptiSparseMET Designs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

# 1. Purpose

Optimisation identifies a design that is best under a stated objective and a
stated set of assumptions. Validation asks a different question: does that
advantage persist when the assumptions, resources, or realised data change?
Therefore, a defensible multi-environment trial (MET) recommendation should
establish: (i) recovery of a known structure under simulation, (ii)
out-of-sample performance with historical response evidence, (iii) superiority
or efficiency relative to explicit reference designs, (iv) robustness to
covariance, variance, missing-data, and site-loss perturbations, and (v)
stability of the released breeding decisions.

This vignette provides a reproducible synthetic demonstration. It does not
claim empirical superiority for a particular crop or breeding programme.
Programme-specific claims require historical multi-year data and prospective
field validation.

# 2. Benchmarking inputs

The environmental and design benchmarks use the same objects as the production
workflow.

| Input | Status | Format | Essential identifiers |
|---|---|---|---|
| Environmental kernels | Required for environmental-model comparison | Named list of square numeric matrices | Identical environment row and column names |
| Genetic relationship matrix (`G`) | Required for design comparison | Square numeric matrix | Treatment row and column names |
| Allocation designs | Required for design comparison | Named list of treatment-by-environment matrices | Treatment rows; environment columns |
| Environmental covariance (`Sigma_E`) | Optional central structure | Named square numeric matrix | Environment row and column names |
| Historical MET responses | Optional response evidence | Long data frame | Genotype, environment, value; year is recommended |
| Dedicated interaction hypotheses | Optional, pre-specified structural evidence | Long data frame | `interaction`, `parent`, `modality`, `variable` |
| Actual replication (`reps`) | Optional, strongly recommended after fieldbook construction | Treatment-by-environment numeric matrix | Same dimensions as allocation |
| Plot cost and capacity | Optional operational evidence | Scalar or environment-named vectors | Environment names |
| Seed inventory | Optional operational constraint | Treatment-named vector or two-column data frame | `Treatment`, `SeedAvailable` |
| Seed required per plot | Conditional on seed inventory | Scalar, environment-named vector, or treatment-by-environment matrix | Treatment and environment names when matrix-valued |

Historical response values should be adjusted genotype estimates, best linear
unbiased estimators (BLUEs), or best linear unbiased predictors (BLUPs) derived
under a coherent model. Raw plot values should first be analysed for block,
spatial, and trial-design effects. Without historical MET responses, the
package does not invent modality weights: the central covariance remains
independent and the environmental kernels are evaluated as separate structural
scenarios.

## 2.1 Reference rows and dimensions

Historical responses use one genotype--environment--year record per adjusted
value:

| genotype | environment | year | adjusted_value |
|---|---|---|---:|
| G001 | E1 | 2023 | 4.82 |
| G001 | E2 | 2023 | 4.31 |
| G002 | E1 | 2023 | 5.07 |

The first three columns are identifiers; `adjusted_value` is numeric. If
several traits are analysed, benchmark them separately or supply the chosen
selection index as the value column.

Seed inventory contains one network-wide quantity per treatment:

| Treatment | SeedAvailable |
|---|---:|
| G001 | 185 |
| G002 | 92 |
| G003 | 240 |

An environmental block has one row per environment. Numeric weather and soil
variables can be combined with categorical management variables:

| environment | mean_temperature | rainfall | soil_pH | irrigation |
|---|---:|---:|---:|---|
| E1 | 24.6 | 512 | 5.8 | supplemental |
| E2 | 26.1 | 438 | 6.2 | rainfed |
| E3 | 23.9 | 605 | 5.5 | supplemental |

A dedicated interaction specification contains one row per parent variable:

| interaction | parent | modality | variable |
|---|---|---|---|
| temperature_x_density | temperature | weather | mean_temperature |
| temperature_x_density | density | soil | bulk_density |

The specification is a pre-defined hypothesis list, not a response-derived
screen. `build_variable_interaction_kernels()` verifies that every column
survived quality control and reports tensor dimension, residual information,
effective kernel rank, bandwidths, and strong-heredity parent kernels.

Allocation and replication objects are matrices rather than data frames. Rows
are treatment IDs and columns are environment IDs. Allocation is binary;
replication is zero for an absent cell and a positive plot count for a present
cell.

```{r benchmark-input-shape, eval=FALSE}
allocation <- matrix(
  c(1, 1, 0,
    1, 0, 1,
    0, 1, 1),
  nrow = 3, byrow = TRUE,
  dimnames = list(c("G001", "G002", "G003"), c("E1", "E2", "E3"))
)
reps <- allocation
reps["G001", "E2"] <- 2

designs <- list(
  proposed = list(allocation_matrix = allocation, reps = reps),
  comparator = comparator_allocation
)
```

# 3. Known-truth environmental covariance recovery

Known-truth simulation tests whether the calibration can recover a covariance
constructed from known environmental components. In this example, weather,
soil, and independent variation contribute 65%, 25%, and 10%, respectively.

```{r known-truth-environment, eval=FALSE}
library(OptiSparseMET)

envs <- paste0("E", 1:6)

K_weather <- outer(
  1:6, 1:6,
  function(i, j) exp(-abs(i - j) / 1.5)
)
K_soil <- outer(
  c(1, 3, 2, 6, 4, 5),
  c(1, 3, 2, 6, 4, 5),
  function(i, j) exp(-abs(i - j) / 1.5)
)
dimnames(K_weather) <- dimnames(K_soil) <- list(envs, envs)

environment_recovery <- simulate_environment_model_benchmark(
  kernels = list(weather = K_weather, soil = K_soil),
  true_weights = c(
    weather = 0.65,
    soil = 0.25,
    identity = 0.10
  ),
  n_genotypes = 250,
  n_years = 10,
  sigma_g2 = 1,
  sigma_e2 = 1,
  reps_per_cell = 2,
  missing_rate = 0.05,
  n_boot = 500,
  seed = 2026
)

environment_recovery$weight_recovery
environment_recovery$benchmark$scores
environment_recovery$benchmark$cross_validation
```

Four quantities should be examined. First, the estimated component weights
should approach the known values as the number of genotypes and years
increases. Second, the historically calibrated covariance should have low
off-diagonal root-mean-square error (RMSE) and relative Frobenius error. Third,
leave-year-out and leave-environment-out errors should remain acceptable.
Fourth, bootstrap weight distributions should reveal whether apparently precise
weights are genuinely identifiable.

The `single_equal_weight` row is deliberately included as a comparator. It
does not become the package default. If equal weighting performs well in one
simulation, that result supports that simulated structure only.

# 4. Retrospective validation with historical MET responses

When historical data are available, blocked validation should respect the
deployment problem. A new year is predicted by leaving an entire year out; a
new location is challenged by leaving an environment out.

```{r retrospective-environment, eval=FALSE}
environment_validation <- benchmark_environment_models(
  kernels = environmental_kernels,
  historical = historical_met,
  genotype_col = "genotype",
  environment_col = "environment",
  value_col = "adjusted_value",
  year_col = "year",
  interaction_policy = "evidence",
  n_boot = 500,
  seed = 2026
)

environment_validation$scores
environment_validation$cross_validation
environment_validation$interaction_evidence
```

The in-sample covariance score and blocked prediction error must be reported
together. A flexible model can reproduce an empirical matrix yet fail on a
held-out year. Moreover, interaction activation is tested against the
main-only model; an interaction kernel remains a structural uncertainty
scenario when response evidence is insufficient.

Dedicated variable interactions should first be assessed separately:

```{r retrospective-variable-interactions, eval=FALSE}
variable_evidence <- assess_variable_interactions(
  kernels = environmental_kernels,
  historical = historical_met,
  genotype_col = "genotype",
  environment_col = "environment",
  value_col = "adjusted_value",
  year_col = "year",
  alpha = 0.05,
  p_adjust = "holm",
  n_boot = 500,
  seed = 2026
)

central_environment_model <- calibrate_environment_covariance(
  kernels = environmental_kernels,
  historical = historical_met,
  genotype_col = "genotype",
  environment_col = "environment",
  value_col = "adjusted_value",
  year_col = "year",
  eligible_interactions = variable_evidence$supported,
  interaction_policy = "evidence",
  n_boot = 500,
  seed = 2026
)

variable_evidence$evidence
central_environment_model$interaction_evidence
```

Component-wise support requires positive blocked-validation improvement, a
Holm-adjusted one-sided paired sign test below the declared threshold, and a
positive fitted covariance contribution. The aggregate calibration then tests
the supported set jointly. This two-stage rule limits interaction multiplicity
while retaining every screened kernel as a structural sensitivity candidate.
Bootstrap inclusion should be interpreted as weight-selection stability, not
as causal evidence.

# 5. Reference designs at equal sparse budget

The design comparator constructs: (i) complete testing as an information upper
bound, (ii) naive independent random sampling, (iii) M3-inspired
coverage-first sparse allocation, and (iv) strict M4 equireplication when the
requested degree sequence is realisable.

```{r reference-designs, eval=FALSE}
treatments <- paste0("L", 1:52)
environments <- paste0("E", 1:4)
common <- treatments[1:4]

reference_designs <- make_met_benchmark_designs(
  treatments = treatments,
  environments = environments,
  n_test_entries_per_environment = 28,
  target_replications = 2,
  common_treatments = common,
  m4_fallback = "omit",
  seed = 2026
)

reference_designs$diagnostics
reference_designs$m4_feasibility
```

The complete design is not forced to satisfy the sparse budget. It defines the
maximum-information reference and is subsequently marked infeasible if it
exceeds plots or seed. Naive random sampling can leave treatments untested; the
coverage metric makes that failure explicit. An infeasible M4 is omitted or
labelled `M4_approximate`; it is never presented as exact equireplication.

# 6. Paired performance and robustness benchmark

The following example evaluates all designs under two covariance structures,
three residual variances, and loss of one site. The same genetic and residual
standard-normal draws are used for every design within a scenario.

```{r paired-design-benchmark, eval=FALSE}
set.seed(2026)
marker_surrogate <- matrix(rnorm(52 * 80), 52, 80)
G <- tcrossprod(marker_surrogate) / 80 + diag(52) * 0.20
dimnames(G) <- list(treatments, treatments)

Sigma_central <- matrix(0.35, 4, 4)
diag(Sigma_central) <- 1
dimnames(Sigma_central) <- list(environments, environments)

Sigma_weak <- matrix(0.10, 4, 4)
diag(Sigma_weak) <- 1
dimnames(Sigma_weak) <- list(environments, environments)

seed_available <- setNames(
  ifelse(seq_along(treatments) <= 16, 5, 3),
  treatments
)

family_info <- data.frame(
  Treatment = treatments,
  Family = rep(paste0("F", 1:4), length.out = length(treatments))
)
seed_info <- data.frame(
  Treatment = treatments,
  SeedAvailable = unname(seed_available)
)

optimised_common <- optimize_common_treatments(
  G = G,
  Sigma_E = Sigma_central,
  treatment_info = family_info,
  seed_info = seed_info,
  seed_required_per_plot = setNames(rep(1, 4), environments),
  entry_capacities = setNames(rep(28L, 4), environments),
  n_groups = 4,
  pair_aggregate = "maximin",
  seed = 2026
)

optimised_allocation <- allocate_sparse_met(
  treatments = treatments,
  environments = environments,
  allocation_method = "random_balanced",
  n_test_entries_per_environment = 28,
  target_replications = 2,
  common_treatments = optimised_common$selected,
  seed_available = seed_available,
  seed_required_per_environment = setNames(rep(1, 4), environments),
  balance = "both",
  balance_iter = 5000,
  Sigma_E = Sigma_central,
  pair_aggregate = "maximin",
  allocation_group_source = "Family",
  treatment_info = family_info,
  min_env_per_group = 2,
  seed = 2026
)
optimised_reps <- optimised_allocation$allocation_matrix
reference_designs$designs$OptiSparseMET_optimised <- list(
  allocation_matrix = optimised_allocation$allocation_matrix,
  reps = optimised_reps
)

design_benchmark <- benchmark_met_designs(
  designs = reference_designs,
  G = G,
  Sigma_E = Sigma_central,
  Sigma_E_candidates = list(
    central = Sigma_central,
    weakly_correlated = Sigma_weak
  ),
  sigma_g2 = 1,
  sigma_e2 = c(0.5, 1, 2),
  environment_dropout = list(
    none = character(),
    fourth_site_lost = "E4"
  ),
  cost_per_plot = c(E1 = 1.0, E2 = 1.0, E3 = 1.2, E4 = 1.3),
  fixed_plot_overhead = c(E1 = 8, E2 = 8, E3 = 8, E4 = 8),
  environment_plot_capacity = c(E1 = 36, E2 = 36, E3 = 36, E4 = 36),
  seed_available = seed_available,
  seed_required_per_plot = 1,
  minimum_seed_buffer = 0,
  n_sim = 500,
  select_fraction = 0.10,
  seed = 2026
)

design_benchmark$design_summary
design_benchmark$scenario_results
design_benchmark$feasibility
optimised_common$comparison
optimised_common$selected
```

The `OptiSparseMET_optimised` row is now compared with every explicit reference
under the same realised genetic effects and errors. The principal result is not
the largest single accuracy. A release candidate should satisfy four
conditions: (i) it is plot- and seed-feasible, (ii) every treatment is observed
at least once unless exclusion is an explicit breeding decision, (iii) its
average performance is competitive, and (iv) its worst-scenario performance is
acceptable. Scenario probabilities should be supplied only when they are
empirically defensible. Otherwise, the uniform summary is a sensitivity
analysis rather than a Bayesian prior.

# 7. Missing environmental covariates

Environmental-data robustness is assessed by masking covariate cells, rebuilding
the full quality-control and imputation workflow, and rescoring the covariance
candidates. The zero-missingness analysis is the reference.

```{r environmental-missingness, eval=FALSE}
missingness_benchmark <- benchmark_environment_missingness(
  weather = weather_covariates,
  soil = soil_covariates,
  management = management_covariates,
  geography = geography_covariates,
  missing_rates = c(0, 0.05, 0.10, 0.20),
  n_repeats = 50,
  historical = historical_met,
  build_args = list(
    missing_action = "impute",
    impute = "median",
    include_interactions = TRUE,
    interaction_mode = "anova",
    variable_interactions = dedicated_interaction_hypotheses
  ),
  benchmark_args = list(
    genotype_col = "genotype",
    environment_col = "environment",
    value_col = "adjusted_value",
    year_col = "year",
    interaction_policy = "evidence",
    n_boot = 0
  ),
  seed = 2026
)

missingness_benchmark$summary
missingness_benchmark$masking
missingness_benchmark$warnings
```

This experiment evaluates missing-completely-at-random covariates. It does not
cover systematic sensor failure during extreme weather, preferentially missing
management records, or other informative mechanisms. Those mechanisms should
be encoded as additional programme-specific perturbations. Low-coverage
quality-control warnings created by deliberate masking are retained in
`warnings` and counted in `masking`; they are not discarded or emitted
repeatedly during the benchmark.

# 8. Cost--performance Pareto frontier

The Pareto plot uses one statistical endpoint. The bottom axis measures
normalised cost minimisation from 0% to 100%; the upper axis reports actual cost
from the maximum to the minimum observed value.

```{r pareto, eval=FALSE}
plot_pareto_frontier(
  design_benchmark,
  x = "cost",
  metric = "accuracy",
  label_designs = TRUE
)
```

For example, if the observed costs range from 10 to 1, cost 10 is aligned with
0% cost minimisation and cost 1 with 100%. Accuracy is shown directly as a
percentage. Genetic gain can instead be selected with `metric = "gain"` and is
then expressed as a percentage of the maximum observed gain. The endpoints are
not overlaid because they answer different breeding questions. A design marked
infeasible cannot define the frontier: it remains visible as a grey cross,
whereas feasible dominated designs remain grey circles.

# 9. Stability of released decisions

Repeated perturbations may change the selected common treatments, recommended
site capacities, or allocation cells. Stability should therefore be reported
at the decision level, not only at the objective level.

```{r stability, eval=FALSE}
stability <- summarize_design_stability(
  common_sets = list(
    bootstrap_1 = common_result_1$selected,
    bootstrap_2 = common_result_2$selected,
    bootstrap_3 = common_result_3$selected
  ),
  site_capacities = list(
    bootstrap_1 = capacity_result_1,
    bootstrap_2 = capacity_result_2,
    bootstrap_3 = capacity_result_3
  ),
  allocations = list(
    bootstrap_1 = allocation_result_1,
    bootstrap_2 = allocation_result_2,
    bootstrap_3 = allocation_result_3
  )
)

stability$common_frequency
stability$common_jaccard
stability$capacity_summary
stability$allocation_frequency
```

Treatment selection frequency identifies stable anchors. Pairwise Jaccard
coefficients quantify whole-set agreement. Capacity medians, interquartile
ranges, and extrema show whether a recommendation depends on a narrow
assumption. Hence, an unstable nominal optimum can be replaced by a slightly
less efficient but more reproducible design before field release.

# 10. Minimum evidence for a strong claim

A quantitative-genetics report should provide:

1. The exact comparator designs and their budgets.
2. The environmental covariance candidates and variance-component grid.
3. Leave-year-out and leave-environment-out results when historical responses
   exist.
4. Paired Monte Carlo means, standard errors, and 95% intervals.
5. Plot, seed, coverage, and site-capacity feasibility.
6. Average and worst-scenario accuracy, gain, PEV, and CDmean.
7. Missing-covariate and complete-site-loss stress tests.
8. Common-set, capacity, and allocation stability.
9. A discrete, feasibility-constrained, non-dominated cost--performance
   frontier.

These analyses do not guarantee future field performance. However, they make
the assumptions, comparisons, and operational compromises auditable. That is
the appropriate standard for converting an optimised sparse MET design into a
breeding decision.
