---
title: "OptiSparseMET: Robust Design of Sparse Multi-Environment Trials"
output: rmarkdown::html_vignette
header-includes:
  - |
    <script>
    window.MathJax = { tex: { inlineMath: [['\\(','\\)']], displayMath: [['\\[','\\]']] }, svg: { fontCache: 'global' } };
    </script>
    <script defer src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
vignette: >
  %\VignetteIndexEntry{OptiSparseMET: Robust Design of Sparse Multi-Environment Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

---

# 1. Introduction

## 1.1 The core problem

A multi-environment trial (MET) evaluates candidate genotypes across the
target population of environments (TPE), defined by locations, years, or
location-by-year combinations, to support selection for broad or specific
adaptation. In modern breeding programmes, however, the number of
candidates $J$ usually exceeds the plot capacity of any one environment.
Complete testing of all $J$ genotypes in all $I$ environments is therefore
infeasible.

Sparse testing addresses this constraint by observing only a planned subset of
genotype-by-environment cells. The scientific problem is not simply to omit
cells, but to determine which omissions preserve information about genetic
merit and genotype-by-environment interaction (G×E). The operational problem is
simultaneously to respect local field geometry, heterogeneous site capacity,
and a finite network-wide seed inventory.

Accordingly, an effective sparse MET must achieve three objectives:
(i) retain direct and model-assisted genetic connectivity; (ii) allocate
replication where it contributes most to prediction and selection; and
(iii) produce field books that can be planted without exceeding plots or seed.

## 1.2 What OptiSparseMET does

`OptiSparseMET` addresses sparse MET design at two linked levels:

1. **Across-environment allocation** — which lines go to which environments,
   how many times each line is replicated across the trial, and whether the
   resulting connectivity is sufficient for cross-environment inference.

2. **Within-environment design** — how the lines assigned to each environment
   are arranged in the field, how checks are distributed, how blocks control
   local heterogeneity, and how spatial gradients are managed.

The package embeds these levels within an environmental evidence framework.
Weather, soil, management, and geographic similarities are retained as
separate main-effect kernels, while weather-by-soil,
weather-by-management, and soil-by-management kernels represent explicit
cross-modality hypotheses. Historical MET responses, when available, calibrate
genetic environment covariance. Without such responses, no environmental
weights are requested: the identity matrix defines the central covariance, and
the separate kernels define unweighted maximin sensitivity scenarios.

The two design levels are therefore linked through (i) the statistical model,
(ii) the information matrix that governs prediction precision, and (iii) the
physical constraints that determine which local layouts can be implemented.

### How the two levels are linked

**Link 1 — The incidence matrix couples allocation to estimation.**
For genotype-by-environment effects $u$ with prior covariance
$\sigma_g^2(\Sigma_E\otimes G)$, `met_information()` uses the absorbed
random-effect coefficient matrix

$$C_{uu}=D-DX(X^\top D X)^{-1}X^\top D+
\{\sigma_g^2(\Sigma_E\otimes G)\}^{-1}.$$

Here $X$ represents fixed environment means and $D$ contains the information
from the observed genotype-by-environment cells. Allocation determines which
diagonal entries of $D$ are non-zero; actual replication, residual variance,
and within-environment efficiency determine their magnitude. The prior
precision couples environments through $\Sigma_E$ and lines through $G$.
Prediction error comes from $C_{uu}^{-1}$, so both design levels affect the
selection criterion.

**Link 2 — Allocation fixes which within-environment designs are feasible.**
With allocation incidence $M$, the number of distinct lines assigned to
environment $e$ is $k_e = \sum_j M_{je}$. Once allocation is complete, $k_e$
is fixed. The within-environment design must then arrange exactly $k_e$
treatments across the available $n_{\text{rows}} \times n_{\text{cols}}$
plots. If $k_e$ is incompatible with the blocking structure — for example,
if $k_e$ is not a multiple of the block size, or if $k_e$ exceeds field
capacity — the within-environment design is infeasible regardless of how good
the allocation was. **Allocation and field geometry must be co-designed.**

**Link 3 — Block structure affects cross-environment inference.**
Consider two environments sharing $n_{\text{shared}}$ lines. The precision of
the cross-environment genetic correlation estimate depends on how well those
shared lines are estimated within each environment, which in turn depends on
the block efficiency within each environment. Formally, the effective
replication of a shared line $j$ across environments is modulated by the
efficiency factor $\epsilon_e$ of the design in environment $e$. In the
package's MET information model, its observation-information contribution is:

$$d_{je} = \frac{r_{je}\epsilon_e}{\sigma_e^2}$$

where $r_{je}$ is the number of plots for line $j$ in environment $e$ and
$\epsilon_e \in (0, 1]$ is the efficiency factor relative to the chosen
reference design. A poor within-environment design (low $\epsilon_e$) reduces
information and increases prediction error, which degrades G×E inference even
when the allocation incidence
structure is ideal. **The block design within each environment affects the
quality of cross-environment comparisons.**

**Link 4 — The mean coefficient of determination (CDmean) depends on both
levels simultaneously.**
For the chosen genetic target $h_j$ (for example, the across-TPE mean), the
line-specific coefficient of determination is

$$CD_j = 1-\frac{\text{PEV}(h_j)}{\text{Var}(h_j)}, \qquad
\text{CDmean}=\frac{1}{J}\sum_{j=1}^{J}CD_j.$$

The prior variance in the denominator respects both the relationship diagonal
and the selected environment target. Prediction-error variance (PEV) depends
on allocation and actual
cell replication through the absorbed mixed-model information matrix, and on
within-environment efficiency. Neither design level can substitute for the
other.

Standard MET tools address one level at a time: either choosing an allocation
strategy without regard to field feasibility, or constructing within-environment
designs without regard to how many lines the allocation will deliver. This
leads to inconsistencies — for example, an allocation that requires a
non-integer number of blocks per replicate, or a field design that assumes more
lines than the allocation provides. `OptiSparseMET` makes both levels explicit
and coordinates them: the allocation output specifies exactly which lines enter
each environment, and the within-environment engine receives precisely that set,
ensuring that both the incidence structure and the blocking structure are
optimised consistently within the same statistical framework.

---

# 2. Statistical Framework

## 2.1 The linear mixed model

Phenotypic observations in plant breeding trials are analysed using a linear
mixed model. For a single environment, a typical form is:

$$y = X\beta + Zg + e$$

where:

- $y$ is the vector of observed phenotypic values (one per plot)
- $X\beta$ captures fixed effects: the overall mean, environment effects,
  check effects, and any fixed treatment contrasts
- $Zg$ captures random line effects: $g \sim N(0,\, K\sigma_g^2)$, where
  $K$ is a genomic relationship matrix (GRM) or pedigree numerator relationship
  matrix (A), and $\sigma_g^2$
  is the genetic variance
- $e \sim N(0,\, R\sigma_e^2)$ is the vector of residuals, with $R$
  encoding the spatial covariance structure

The design matrix $Z$ — whose columns correspond to lines and whose rows
correspond to plots — is determined entirely by the experimental design.
**A good design is one that makes $Z$ well-conditioned for estimation.**

## 2.2 Why design affects inference

When $Z$ is sparse, not all lines are observed in all environments.
Cross-environment inference then depends on either:

- **Design-based connectivity**: lines that co-occur in the same environment
  create direct comparisons; common treatments that appear in every environment
  create bridges between sites.
- **Model-based connectivity**: the relationship matrix $K$ propagates
  information between genetically related lines, even if they never appear in
  the same environment.

Both mechanisms are important, but model-based connectivity is less robust
because it depends on the assumed covariance structure being correct. A
well-designed sparse MET builds in design-based connectivity as a safeguard.

## 2.3 Efficiency criteria

Given a design, its statistical efficiency can be measured by how well it
supports estimation and prediction of genetic values. The three main criteria
used by `OptiSparseMET` are:

**A-optimality** minimises the average variance of treatment comparison
estimates (contrasts). Formally it minimises $\text{trace}(C^{-1})$ where
$C$ is the coefficient matrix of the treatment effects. A lower value is
better. Its reciprocal, **A-efficiency** $= 1 / A_{\text{criterion}}$, is
sometimes more interpretable because larger values are better.

**D-optimality** maximises the determinant of the estimable information matrix,
which minimises generalised variance and the volume of the confidence
ellipsoid for treatment contrasts. It is more sensitive to the overall spread
of information across all contrasts than A-optimality, which focuses on the
average. Again, **D-efficiency**
$= 1 / D_{\text{criterion}}$ is used so larger values are better.

**CDmean** is the criterion most directly tied to genomic prediction. For
target genetic value $h_j$, it is defined as:

$$\text{CDmean}=\frac{1}{J}\sum_{j=1}^{J}
\left\{1-\frac{\text{PEV}(h_j)}{\text{Var}(h_j)}\right\}.$$

The line-specific prior variance can differ through the relationship diagonal
and the environment target. The package bounds numerical reliability estimates
to $[0,1]$; values close to 1 indicate precise prediction. CDmean incorporates
both the relationship model and design information.

> **For novice readers**: Think of CDmean as "what fraction of the true
> genetic value the design allows you to recover on average." A CDmean of 0.85
> means the design, on average, captures 85% of the genetic signal.

---

# 3. Two-Level Design Structure

## 3.1 Across-environment allocation

The resource identity underlying balanced sparse allocation is:

$$N = J \times r = \sum_{e=1}^{I} k_e$$

where:

| Symbol | Meaning |
|--------|---------|
| $N$ | Total number of line-by-environment observations |
| $J$ | Total number of candidate lines |
| $r$ | Number of environments each line enters (replication depth) |
| $I$ | Number of environments |
| $k_e$ | Number of lines tested in environment $e$ (coverage breadth) |

Given a fixed total resource $N$, this identity formalizes a fundamental
tradeoff: **increasing coverage breadth** (testing more lines per environment,
larger $k_e$) **necessarily reduces replication depth** (each line appears in
fewer environments, smaller $r$). Equal site sizes are the special case
$\sum_e k_e = I k$. No design can simultaneously maximise both.

> **For novice readers**: With 100 lines and 4 environments, 400
> line-by-environment slots allow all 100 lines in all 4 environments
> ($r=4$, $k=100$). With 200 slots, testing 50 lines per environment allows
> each line to appear in 2 environments ($r=2$, $k=50$). With 100 slots,
> testing 25 lines per environment allows each line to appear once
> ($r=1$, $k=25$). In each equal-site example, $Jr=Ik$.

The practical constraint is that $k$ is bounded by the number of plots
available per environment, and $r$ is bounded by seed availability. The
allocation problem is to choose which lines go to which environments, subject
to these bounds.

When common treatments are included, a number $C$ of lines are assigned to
**every** environment before sparse allocation begins. Their count is subtracted
from each environment's capacity:

$$k_e^* = k_e - C$$

The sparse-allocatable lines number $J^* = J - C$, and the modified resource
identity applies to them:

$$J^* \times r = \sum_{e=1}^{I} k_e^*$$

For equal environment sizes this reduces to $J^*r=Ik^*$.

## 3.2 Within-environment design

Each environment receives an independent field design constructed by one of
two engines:

- `met_prep_famoptg()` for block-based repeated-check designs, including
  augmented, partially replicated (p-rep), and randomised complete block
  design (RCBD)-type layouts
- `met_alpha_rc_stream()` for row-column alpha designs suited to large-scale
  spatial field structure

The choice between them depends on field geometry, check structure,
replication goals, and the spatial model anticipated at analysis. They can
differ across environments within the same trial.

---

# 4. Pipeline Overview

The `OptiSparseMET` workflow proceeds in four stages, with an optional
shortcut that runs all four together.

```
STAGE 0   Verify per-environment capacity
          suggest_safe_k()  OR  min_k_for_full_coverage()
          |
STAGE 1   Allocate lines across environments
          allocate_sparse_met()
          |
STAGE 2   Reconcile allocation and replication against one shared inventory
          plan_sparse_met_design()
          (assign_replication_by_seed() remains the local-site helper)
          |
STAGE 3   Build within-environment field designs
          met_prep_famoptg()  OR  met_alpha_rc_stream()
          (optional: evaluate and optimise each design)
          |
STAGE 4   Assemble the combined MET field book
          combine_met_fieldbooks()

          --- OR: run stages 1-4 together ---

          plan_sparse_met_design()
```

```{r pipeline-diagram, eval=FALSE}
library(DiagrammeR)

grViz("
digraph OptiSparseMET {
  graph [layout = dot, rankdir = LR]
  node [shape = rectangle, style = rounded, fontname = Helvetica]

  A [label = 'Input\nLines  Environments\nFamilies / GRM / A\nSeed inventory']
  B [label = 'Stage 0: Verify capacity\nsuggest_safe_k()']
  C [label = 'Stage 1: Allocation\nallocate_sparse_met()\n(M3 or M4)']
  D [label = 'Stage 2: Network seed ledger\nplan_sparse_met_design()']
  E [label = 'Stage 3: Field design\nmet_prep_famoptg()\nor met_alpha_rc_stream()']
  F [label = 'Stage 4: Assemble\ncombine_met_fieldbooks()']
  G [label = 'Outputs\ncombined_field_book\nenvironment_summary\nefficiency_summary']

  A -> B -> C -> D -> E -> F -> G
}
")
```

---

# 5. Feasibility Helpers

Before any allocation, it is essential to verify that the chosen
per-environment capacity $k$ is large enough to assign every non-common line at
least once. Passing a capacity that is too small causes `allocate_sparse_met()`
to stop with an informative error.

The minimum capacity for full coverage under equal environment sizes is:

$$k_{\min} = \left\lceil \frac{J^*}{I} \right\rceil + C$$

where $\lceil \cdot \rceil$ is the ceiling function. This is the floor below
which any sparse allocation must leave some lines unassigned.

`OptiSparseMET` provides three helpers:

| Function | Action |
|----------|--------|
| `min_k_for_full_coverage()` | Returns $k_{\min}$ exactly |
| `suggest_safe_k()` | Returns $k_{\min}$ plus a user-defined buffer |
| `warn_if_k_too_small()` | Emits a non-fatal warning if $k < k_{\min}$ |

Always call one of these before `allocate_sparse_met()`.

```{r feasibility, eval=FALSE}
library(OptiSparseMET)

data("OptiSparseMET_example_data")
x <- OptiSparseMET_example_data

# Strict arithmetic minimum
min_k_for_full_coverage(
  n_treatments_total  = length(x$treatments),    # 120
  n_environments      = length(x$environments),  # 4
  n_common_treatments = length(x$common_treatments)  # 8; J*=112
)
# min_sparse_slots_per_environment = 28  (= ceil(112/4))
# min_total_entries_per_environment = 36 (= 28 + 8)

# Safe value with buffer of 3
suggest_safe_k(
  treatments        = x$treatments,
  environments      = x$environments,
  common_treatments = x$common_treatments,
  buffer            = 3
)
# Returns 39

# Non-fatal check (useful inside scripts)
warn_if_k_too_small(
  treatments                     = x$treatments,
  environments                   = x$environments,
  n_test_entries_per_environment = 39,
  common_treatments              = x$common_treatments
)
# No warning: 4 * (39 - 8) = 124 >= 112 sparse lines
```

---

## 5.1 Input contract: required and optional data

This section defines the inputs expected by the end-to-end package workflow.
A required input has no usable default. A
conditional input becomes required only when its associated analysis is
requested. Optional evidence can improve a design, but its absence must not be
hidden by invented values.

### 5.1.1 General formatting rules

1. Treatment and environment identifiers must be non-missing, non-empty, and
   unique within their defining table.
2. The same identifier must be written identically in every object. Matching is
   case-sensitive.
3. Check treatments are supplied to the local field-design specifications; they
   are not included in the `treatments` vector used for sparse allocation.
4. Named vectors are recommended whenever values differ among environments or
   treatments. Names, rather than position, then control alignment.
5. Relationship and covariance matrices must be numeric, finite, square,
   symmetric, and identically named by row and column.
6. Seed availability and seed requirement must use the same unit; grams are a
   common choice. An absent treatment-by-environment cell consumes 0 g.
7. Missing environmental observations must remain `NA` until an explicit
   quality-control and imputation rule is applied.

### 5.1.2 End-to-end design: `plan_sparse_met_design()`

#### Required arguments

| Argument | Accepted format | Requirement |
|---|---|---|
| `treatments` | Unique character vector | Test-entry IDs. Do not include local checks. |
| `environments` | Unique character vector of length at least 2 | Environment IDs used consistently throughout the design. |
| `n_test_entries_per_environment` | Positive integer scalar, positional integer vector, or named integer vector | Number of distinct test entries at each environment, including common entries but excluding local check plots and extra replication plots. |
| `env_design_specs` | Named list with one list per environment | Local design engine, field dimensions, checks, and engine-specific settings. Names must cover every value in `environments`. |

Example core vectors:

```{r input-core-vectors, eval=FALSE}
treatments <- sprintf("G%03d", seq_len(120))
environments <- c("Environment_A", "Environment_B", "Environment_C")
n_test_entries_per_environment <- c(
  Environment_A = 80L,
  Environment_B = 60L,
  Environment_C = 60L
)
```

#### Optional and conditional arguments

| Argument | Default | When it is used |
|---|---:|---|
| `allocation_method` | `"random_balanced"` | Use `"equireplicate"` when equal environment replication and the required degree sequence are feasible. `"M3"` and `"M4"` remain aliases. |
| `target_replications` | Inferred | Target number of environments per non-common treatment; strict for `equireplicate`. |
| `common_treatments` | `NULL` | Character vector of preselected entries. When supplied, these identities are fixed and required at every environment; their local replication may differ. |
| `balance` | `"none"` | `"env_pair"`, `"line_pair"`, or `"both"` for concurrence refinement. |
| `balance_iter`, `balance_seed` | `2000L`, `NULL` | Iteration budget and reproducibility seed for concurrence refinement. |
| `treatment_info` | `NULL` | Required for family-guided allocation; also supplies family labels to local designs. |
| `seed_info` | `NULL` | Activates network-wide seed accounting and seed-aware local replication. |
| `seed_required_per_plot` | `NULL` | Required with `seed_info`; scalar, named vector/list, or an environment-level data frame. |
| `minimum_seed_buffer` | `0` | Non-negative reserve retained per treatment after the complete network is planned. |
| `allocation_group_source` | `"none"` | `"Family"`, `"GRM"`, or `"A"` activates genetic grouping. |
| `GRM` | `NULL` | Required when `allocation_group_source = "GRM"`. |
| `A` | `NULL` | Required when `allocation_group_source = "A"`. |
| `id_map` | `NULL` | Required only when treatment IDs differ from relationship-matrix IDs. |
| `group_method` | `"kmeans"` | `"kmeans"` or `"hclust"` for matrix-derived allocation groups. |
| `group_seed`, `group_attempts`, `n_pcs_use` | `1`, `25`, `Inf` | Reproducibility, restarts, and retained principal components for matrix-derived groups. |
| `min_groups_per_environment` | `NULL` | Minimum represented genetic groups per environment, where feasible. |
| `min_env_per_group` | `NULL` | Minimum environments representing each genetic group, where feasible. |
| `balance_groups_across_env` | `TRUE` | Balances group counts across environments. |
| `force_group_connectivity` | `TRUE` | Prioritises groups that have not reached `min_env_per_group`. |
| `allow_approximate` | `FALSE` | Relaxes exact equireplication only for an explicitly exploratory fallback. |
| `seed` | `NULL` | Reproducibility seed for allocation and tie-breaking. |

### 5.1.3 Core tabular objects

#### Treatment information

`treatment_info` is required for family-guided allocation. It must contain one
row per treatment and cover every allocated treatment.

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `Treatment` | character | yes | Identifier matching `treatments`. |
| `Family` | character | yes for family grouping | Breeding family or genetic stratum. |
| Additional columns | any | no | Programme metadata retained outside the core allocator. |

Example:

| Treatment | Family |
|---|---|
| G001 | Fam01 |
| G002 | Fam01 |
| G003 | Fam07 |

#### Network-wide seed inventory

`seed_info` contains one row per treatment. `SeedAvailable` must be finite and
non-negative. In seed-constrained workflows its unit is grams.

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `Treatment` | character | yes | Identifier matching `treatments`. |
| `SeedAvailable` | numeric | yes | Total inventory for the complete network, not inventory per site. |

Example:

| Treatment | SeedAvailable |
|---|---:|
| G001 | 185 |
| G002 | 92 |
| G003 | 240 |

Environment-specific seed requirement may be supplied as a scalar, a named
numeric vector, or this data frame:

| Environment | SeedRequiredPerPlot |
|---|---:|
| Environment_A | 10 |
| Environment_B | 16 |
| Environment_C | 20 |

The allocation debit includes the mandatory local replication. For example,
an alpha design with two replicates at Environment_C requires
\(2\times20=40\) g for
each treatment assigned there.

#### Treatment-to-matrix identifier map

`id_map` is needed only when field treatment IDs and relationship-matrix IDs
differ.

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `Treatment` | character | yes | Field and allocation identifier. |
| `LineID` | character | yes | Corresponding row and column name in `GRM` or `A`. |

Example:

| Treatment | LineID |
|---|---|
| G001 | M001 |
| G002 | M002 |
| G003 | M003 |

### 5.1.4 Relationship and covariance matrices

A genomic relationship matrix (`GRM`) or pedigree numerator relationship
matrix (`A`) uses treatment or mapped line IDs on both axes:

|  | G001 | G002 | G003 |
|---|---:|---:|---:|
| G001 | 1.08 | 0.42 | 0.07 |
| G002 | 0.42 | 1.03 | 0.05 |
| G003 | 0.07 | 0.05 | 0.96 |

An environmental covariance matrix (`Sigma_E`) follows the same rule with
environment IDs on both axes. Additive relationship is the default. A
dominance matrix is optional and should be supplied only when hybrid or
testcross dominance is part of the prediction target.

### 5.1.5 Local field-design specifications

`env_design_specs` is a named list, not a data frame. Each environment must use
one of the two built-in engines.

For `met_prep_famoptg`, the required fields are `design`,
`check_treatments`, `check_families`, `n_blocks`, `n_rows`, and `n_cols`.
Optional pipeline fields include `replication_mode`, `desired_replications`,
`candidate_prep`, `max_prep`, `max_extra_replication_plots`,
`shortage_action`, and a named positive-integer vector
`fixed_treatment_reps`.

For `met_alpha_rc_stream`, the required fields are `design`,
`check_treatments`, `check_families`, `n_reps`, `n_rows`, and `n_cols`.
Allocated entries and their families are inserted by the pipeline. Optional
fields include `min_block_size`, `max_block_size`, `cluster_source`,
`order`, `serpentine`, and `seed`.

All optional `met_prep_famoptg` specification fields are:

| Field or fields | Default | Purpose |
|---|---:|---|
| `replication_mode` | Inferred | Pipeline replication class: `"augmented"`, `"p_rep"`, or `"rcbd_type"`. |
| `desired_replications` | `2L` | Target plots for replicated entries. |
| `candidate_prep` | `NULL` | Entries eligible for partial replication. |
| `max_prep` | `NULL` | Maximum number of partially replicated entries. |
| `max_extra_replication_plots` | `NULL` | Physical ceiling on plots beyond one base plot per allocated entry. |
| `shortage_action` | `"downgrade"` | Seed-shortage response. |
| `fixed_treatment_reps` | Empty named vector | Exact site-specific replication for selected allocated entries. |
| `order`, `serpentine` | `"column"`, `FALSE` | Field traversal direction and alternating traversal. |
| `seed`, `attempts` | `NULL`, `1000` | Randomisation seed and within-block search effort. |
| `warn_and_correct`, `fix_rows` | `TRUE`, `TRUE` | Whether and how an undersized grid is expanded. |
| `cluster_source` | `"Family"` | Adjacency groups from family, `GRM`, or `A`. |
| `GRM`, `A`, `id_map` | `NULL` | Conditional relationship inputs and identifier map. |
| `cluster_method`, `cluster_seed`, `cluster_attempts`, `n_pcs_use` | `"kmeans"`, `1`, `25`, `Inf` | Relationship-based clustering controls. |
| `check_placement`, `check_opt_attempts` | `"random"`, `200` | Random, systematic, or optimised check placement and its search effort. |
| `use_dispersion`, `dispersion_source` | `FALSE`, `"K"` | Optional local dispersion of related entries and the relationship source. |
| `dispersion_radius`, `dispersion_iters`, `dispersion_seed` | `1`, `2000`, `1` | Dispersion neighbourhood, search effort, and seed. |
| `K`, `line_id_map` | `NULL` | Conditional dispersion matrix and its identifier map. |
| `verbose` | `TRUE` | Print the local design summary. |

All optional `met_alpha_rc_stream` specification fields are:

| Field or fields | Default | Purpose |
|---|---:|---|
| `order`, `serpentine` | `"column"`, `FALSE` | Field traversal direction and alternating traversal. |
| `seed`, `attempts` | `NULL`, `5000` | Randomisation seed and within-block search effort. |
| `warn_and_correct`, `fix_rows` | `TRUE`, `TRUE` | Compatibility controls; alpha field geometry remains fixed. |
| `cluster_source` | `"Family"` | Adjacency groups from family, `GRM`, or `A`. |
| `GRM`, `A`, `id_map` | `NULL` | Conditional relationship inputs and identifier map. |
| `cluster_method`, `cluster_seed`, `cluster_attempts`, `n_pcs_use` | `"kmeans"`, `1`, `25`, `Inf` | Relationship-based clustering controls. |
| `n_blocks_per_rep` | `NULL` | Exact incomplete-block count per replicate; otherwise inferred. |
| `min_block_size`, `max_block_size` | `NULL`, `NULL` | Bounds on total block size, including checks. |
| `check_placement`, `check_position_pattern` | `"systematic"`, `"spread"` | Check-position controls. |
| `use_dispersion`, `dispersion_source` | `FALSE`, `"K"` | Optional local dispersion of related entries and the relationship source. |
| `dispersion_radius`, `dispersion_iters`, `dispersion_seed` | `1`, `2000`, `1` | Dispersion neighbourhood, search effort, and seed. |
| `K`, `line_id_map` | `NULL` | Conditional dispersion matrix and its identifier map. |
| `verbose` | `TRUE` | Print the local design summary. |

```{r input-design-specs, eval=FALSE}
env_design_specs <- list(
  Environment_A = list(
    design = "met_prep_famoptg",
    check_treatments = c("CHK_A", "CHK_B"),
    check_families = c("CHECK", "CHECK"),
    n_blocks = 4L,
    n_rows = 12L,
    n_cols = 14L,
    replication_mode = "p_rep",
    desired_replications = 2L,
    max_extra_replication_plots = 36L
  ),
  Environment_B = list(
    design = "met_alpha_rc_stream",
    check_treatments = c("CHK_A", "CHK_B"),
    check_families = c("CHECK", "CHECK"),
    n_reps = 2L,
    n_rows = 10L,
    n_cols = 20L,
    min_block_size = 6L,
    serpentine = TRUE
  ),
  Environment_C = list(
    design = "met_alpha_rc_stream",
    check_treatments = c("CHK_A", "CHK_B"),
    check_families = c("CHECK", "CHECK"),
    n_reps = 2L,
    n_rows = 10L,
    n_cols = 20L,
    min_block_size = 6L,
    serpentine = TRUE
  )
)
```

The physical grid must accommodate the fieldbook produced after check and
replication plots are added. `n_test_entries_per_environment` is therefore not
the same quantity as `n_rows * n_cols`.

### 5.1.6 Environmental evidence

Environmental inputs are optional for constructing a feasible fieldbook but
required for environment-informed grouping, representativeness, or covariance
analysis. Each feature block may be a data frame with an `environment` column
or a matrix with environment row names.

#### Precomputed weather, soil, or management features

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `environment` | character | yes for data frames | Environment identifier. |
| Feature columns | numeric; management may also be categorical | at least one | Audited environmental variables or derived features. |

Weather example:

| environment | T2M_mean_vegetative | PRECTOTCORR_sum_reproductive | VPD_q90 |
|---|---:|---:|---:|
| Environment_A | 26.8 | 318.4 | 1.72 |
| Environment_B | 27.4 | 212.7 | 2.06 |
| Environment_C | 24.1 | 286.3 | 1.48 |

Management example:

| environment | planting_mode | irrigation_mm | nitrogen_kg_ha |
|---|---|---:|---:|
| Environment_A | direct | 420 | 130 |
| Environment_B | direct | 510 | 145 |
| Environment_C | transplanting | 620 | 160 |

Geography requires:

| environment | latitude | longitude |
|---|---:|---:|
| Environment_A | -1.80 | -79.53 |
| Environment_B | 3.93 | -75.02 |
| Environment_C | 3.50 | -76.35 |

#### Historical MET values

Historical values are optional. When supplied, they must be adjusted genetic
values or best linear unbiased estimates, not raw plot observations.

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `genotype` | character | yes | Historical genotype identifier. |
| `environment` | character | yes | Historical environment identifier. |
| `value` | numeric | yes | Adjusted genetic value for the target trait. |
| `year` | integer or character | no | Enables year-specific targets and leave-year-out validation. |

Example:

| genotype | environment | value | year |
|---|---|---:|---:|
| H001 | Environment_A | 0.84 | 2023 |
| H001 | Environment_B | 0.31 | 2023 |
| H002 | Environment_A | -0.12 | 2023 |

Without valid historical MET values, `Sigma_E` is the identity matrix and the
separate environmental kernels remain unweighted maximin scenarios.

### 5.1.7 Lower-level function requirements

When the pipeline functions are called separately:

- `allocate_sparse_met()` requires `treatments`, `environments`, and
  `n_test_entries_per_environment`; `allocation_method` has a default.
- `assign_replication_by_seed()` requires `treatments`, `seed_available`, and
  `seed_required_per_plot`; `replication_mode` defaults to `"augmented"`.
- `met_prep_famoptg()` requires checks, check families, the replicated and
  unreplicated treatment/family vectors, their replication counts, and block
  and field dimensions. Either non-check treatment set may be empty, but not
  both.
- `met_alpha_rc_stream()` requires checks, check families, entry treatments,
  entry families, `n_reps`, `n_rows`, and `n_cols`.

The corresponding function help pages define every advanced optimiser and
efficiency argument.

## 5.2 Minimum working example

The absolute minimum to run the full pipeline from allocation to field book:

```{r min-inputs, eval=FALSE}
library(OptiSparseMET)

# Minimum inputs: just lines, environments, and field dimensions
treatments <- paste0("L", sprintf("%03d", 1:120))
envs       <- c("E1", "E2", "E3", "E4")

# Stage 0: verify k
k <- suggest_safe_k(treatments, envs, buffer = 3)  # 33

# Stage 1: M3 allocation (no common treatments, no grouping)
alloc <- allocate_sparse_met(
  treatments                     = treatments,
  environments                   = envs,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = k,
  seed                           = 1
)

# Stage 2: seed plan (uniform seed, no shortage)
seed_df <- data.frame(
  Treatment     = treatments,
  SeedAvailable = 100L
)
rep_plan <- assign_replication_by_seed(
  treatments             = treatments,
  seed_available         = seed_df,
  seed_required_per_plot = 10L,
  replication_mode       = "augmented"
)

# Stage 3: within-environment design (checks + unreplicated entries)
design <- met_prep_famoptg(
  check_treatments        = c("CHK1", "CHK2"),
  check_families          = c("CHECK", "CHECK"),
  unreplicated_treatments = rep_plan$unreplicated_treatments,
  unreplicated_families   = rep("F1", length(rep_plan$unreplicated_treatments)),
  n_blocks = 4L, n_rows = 10L, n_cols = 12L,
  seed     = 1
)

# Stage 4: combine
met_book <- combine_met_fieldbooks(
  field_books = list(E1 = design$field_book)
)
```


---

# 6. Allocation Strategies

## 6.1 The random balanced strategy (M3)

The `"random_balanced"` method is a coverage-first stochastic allocation
inspired by the M3 method of Montesinos-López et al. (2023). It operates in
two phases:

**Phase 1 (coverage guarantee)**: Every non-common line is assigned to at
least one environment. Lines are processed in random order; each is placed in
the environment with the most remaining capacity, with a preference for
environments where the line's genetic group is not yet represented when
`allocation_group_source` is active.

**Phase 2 (replication filling)**: Remaining slots are filled by repeatedly
selecting lines with the largest replication deficit relative to
`target_replications`, subject to group-balance penalties.

This strategy is appropriate when:
- Environment capacities are heterogeneous
- Exact balanced incomplete block design (BIBD) parameters are not achievable
  for the trial dimensions
- Some flexibility in the incidence structure is acceptable

Unlike the original M3, which allocates environment by environment and may
silently leave some lines with no assignment, this implementation guarantees
full coverage before attempting any additional replication.

## 6.2 The balanced incomplete strategy (M4)

The `"equireplicate"` method implements the M4 allocation of
Montesinos-López et al. (2023). It enforces two structural guarantees:

**Condition 1 — Equal replication**: every non-common line appears in exactly
$r$ environments. This ensures no line is systematically under-evaluated and
is the defining property that distinguishes M4 from M3. Here, replication
means distinct-environment incidence, not repeated plots within a location.
Local plot replication never increases cross-environment connectivity.

**Condition 2 — Exact requested environment sizes**: environment $e$ receives
its requested $k_e^*$ sparse lines. Loads may differ, provided
$\sum_e k_e^* = J^*r$ and the equal-replication degree sequence is realisable.
The familiar $J^*r = Ik^*$ identity is the equal-size special case.

These two conditions are what matter in plant breeding programmes where thousands
of candidate lines are tested across a few environments. Enforcing them means
every line gets the same number of chances to be evaluated — a fundamental
fairness and precision requirement.

> **For novice readers**: Equal replication means that if two breeders each
> select the "top 10% of lines" based on the trial data, they are working from
> the same quality of information for every line. Without equal replication,
> lines tested more often have lower-variance estimates and a systematic
> advantage in selection.

`allow_approximate = FALSE` (the default) enforces both conditions strictly.
The total slot identity and Gale-Ryser degree condition must hold, and the
function stops clearly otherwise. Use `check_equireplicate_feasibility()` before
allocating. The equal-capacity example below reduces to the classic identity:

```{r bibd-check, eval=FALSE}
# With k=39, J*=112, I=4, r=1: available slots = 4*(39-8) = 124,
# required slots = 112*1 = 112. 124 ≠ 112 -> slot identity fails.
check_balanced_incomplete_feasibility(
  n_treatments_total             = 120,
  n_environments                 = 4,
  n_test_entries_per_environment = 39,
  target_replications            = 1,
  n_common_treatments            = 8
)
# feasible   = FALSE  (available = 124, required = 112, difference = 12)
# Adjust k so that J*×r = I×k*: e.g. k=36 -> k*=28, 4*28=112=112*1.

# With k=36: slot identity holds.
check_balanced_incomplete_feasibility(
  n_treatments_total             = 120,
  n_environments                 = 4,
  n_test_entries_per_environment = 36,
  target_replications            = 1,
  n_common_treatments            = 8
)
# feasible = TRUE (difference = 0)
```

Construction uses a greedy load-balanced constructor that guarantees equal replication.

`allow_approximate = TRUE` relaxes the slot identity and accepts minor
replication imbalances. This is a fallback for exploratory use when exact
slot identity cannot be achieved, not the primary path.

## 6.3 M3 versus M4: choosing the right strategy

| Feature | M3 `"random_balanced"` | M4 `"equireplicate"` |
|---------|------------------------|----------------------------|
| Equal replication guarantee | No (approximate) | Yes (`allow_approximate = FALSE`) |
| Exact requested environment sizes | Yes | Yes when the degree sequence is realisable |
| Pairwise co-occurrence | Stochastic | Approximately uniform; compute from `$allocation_matrix %*% t($allocation_matrix)` |
| Handles unequal environment sizes | Yes | Yes, including strict mode |
| Requires total identity $\sum_e k_e^* = J^*r$ | No | Yes (strict mode) |
| Coverage guarantee | Yes (phase 1) | Yes |
| Typical use | Heterogeneous programmes, exploratory | Standard MET networks with equal replication |

In practice, M3 is more flexible and tolerates heterogeneous constraints.
M4 with `allow_approximate = FALSE` (the default) is the correct choice
whenever equal replication is a hard requirement — which it should be in most
standard breeding programmes. Verify the total slot identity and degree sequence
first with `check_equireplicate_feasibility()` and adjust the site loads or $r$
when needed. M4 with `allow_approximate = TRUE` is a fallback when strict equal
replication cannot be satisfied.

```{r allocation-comparison, eval=FALSE}
library(OptiSparseMET)

data("OptiSparseMET_example_data")
x <- OptiSparseMET_example_data

# Verify feasibility first
suggest_safe_k(x$treatments, x$environments,
               common_treatments = x$common_treatments, buffer = 3)
# Returns 39: ceil(112/4) + 8 + 3 = 39

# M3: random balanced allocation
alloc_M3 <- allocate_sparse_met(
  treatments                     = x$treatments,
  environments                   = x$environments,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = 39,
  target_replications            = 1,
  common_treatments              = x$common_treatments,
  seed                           = 123
)

# M4: balanced incomplete allocation with equal replication.
# J*=112, I=4, r=1: need k* = 112/4 = 28, k = 28+8 = 36.
# Slot identity: 112*1 = 4*28 = 112. Verified above.
alloc_M4 <- allocate_sparse_met(
  treatments                     = x$treatments,
  environments                   = x$environments,
  allocation_method              = "equireplicate",
  n_test_entries_per_environment = 36,
  target_replications            = 1,
  common_treatments              = x$common_treatments,
  allow_approximate              = FALSE,
  seed                           = 123
)

alloc_M3$summary  # allocation_method, min/max/mean replication, etc.
alloc_M4$summary  # min_sparse_replication = max_sparse_replication = 1 (equal)
alloc_M4$summary$max_sparse_replication  # equals min_sparse_replication: equal replication confirmed
```

---

## 6.4 Equal-size special case: divisibility by J\*, I, and r

For equal environment sizes, $J^* \times r = I \times k^*$ requires that $J^* \times r$
be exactly divisible by $I$. Whether this is achievable for a given
combination of sparse treatments ($J^*$), environments ($I$), and replication
($r$) depends on the shared factors of these three numbers.

## The divisibility rule

For the slot identity to hold, $I$ must divide $J^* \times r$ exactly.
Every prime factor of $I$ that is absent from $J^*$ must be supplied by $r$.
This has a practical consequence for the most common case in plant breeding:

- **$I = 4$ environments, $J^*$ odd**: $J^* \times 1 = \text{odd}$ (not
  divisible by 4); $J^* \times 2 = 2 \times \text{odd}$ (divisible by 2
  but not by 4 for odd $J^*$); only $r = 4$ guarantees divisibility. But
  $r = 4$ gives $k^* = J^*$ — full replication — which defeats the purpose
  of sparse testing. **Practical fix**: adjust $C$ by 1 so that $J^* = J - C$
  becomes even.

- **$I = 4$ environments, $J^*$ even but not divisible by 4**: $r = 2$
  always works (e.g. $J^* = 110$: $110 \times 2 / 4 = 55$).

- **$I = 3$ environments**: feasibility depends on divisibility by 3. If
  $J^*$ is divisible by 3, any $r$ works. Otherwise $r$ must be a multiple
  of 3.

- **$I = 6$ environments**: requires divisibility by $2 \times 3 = 6$.
  Odd $J^*$ not divisible by 3 requires $r$ divisible by 6.

## Feasibility table: $r = 2$

The table shows $k^*$ when the slot identity holds, and `--` when it does
not for that combination. Add $C$ (common treatments) to $k^*$ to get the
`n_test_entries_per_environment` argument.

| $J^*$ | $I=3$ | $I=4$ | $I=5$ | $I=6$ | $I=7$ | $I=8$ | $I=9$ | $I=10$ |
|------:|------:|------:|------:|------:|------:|------:|------:|-------:|
| 60 | 40 | 30 | 24 | 20 | -- | 15 | -- | 12 |
| 70 | -- | 35 | 28 | -- | 20 | -- | -- | 14 |
| 75 | 50 | -- | 30 | 25 | -- | -- | -- | 15 |
| 80 | -- | 40 | 32 | -- | -- | 20 | -- | 16 |
| 90 | 60 | 45 | 36 | 30 | -- | -- | 20 | 18 |
| 100 | -- | 50 | 40 | -- | -- | 25 | -- | 20 |
| 110 | -- | 55 | 44 | -- | -- | -- | -- | 22 |
| 112 | -- | 56 | -- | -- | 32 | 28 | -- | -- |
| 120 | 80 | 60 | 48 | 40 | -- | 30 | -- | 24 |
| 150 | 100 | 75 | 60 | 50 | -- | -- | -- | 30 |
| 200 | -- | 100 | 80 | -- | -- | 50 | -- | 40 |

## Feasibility table: $r = 3$

| $J^*$ | $I=3$ | $I=4$ | $I=5$ | $I=6$ | $I=7$ | $I=8$ | $I=9$ | $I=10$ |
|------:|------:|------:|------:|------:|------:|------:|------:|-------:|
| 60 | 60 | 45 | 36 | 30 | -- | -- | 20 | 18 |
| 70 | 70 | -- | 42 | 35 | 30 | -- | -- | 21 |
| 75 | 75 | -- | 45 | -- | -- | -- | 25 | -- |
| 80 | 80 | 60 | 48 | 40 | -- | 30 | -- | 24 |
| 90 | 90 | -- | 54 | 45 | -- | -- | 30 | 27 |
| 100 | 100 | 75 | 60 | 50 | -- | -- | -- | 30 |
| 110 | 110 | -- | 66 | 55 | -- | -- | -- | 33 |
| 112 | 112 | 84 | -- | 56 | 48 | 42 | -- | -- |
| 120 | 120 | 90 | 72 | 60 | -- | 45 | 40 | 36 |
| 150 | 150 | -- | 90 | 75 | -- | -- | 50 | 45 |
| 200 | 200 | 150 | 120 | 100 | -- | 75 | -- | 60 |

## What to do when your combination gives `--`

Use `check_equireplicate_feasibility()` to diagnose the problem and
try one of these adjustments:

1. **Adjust $C$ by 1**: adding or removing one common treatment changes $J^*$
   by 1, which may make it divisible by $I$ for the chosen $r$.
2. **Try $r = 2$ instead of $r = 1$**, or $r = 3$ instead of $r = 2$ — the
   extra factor may resolve the divisibility.
3. **Use `random_balanced` (M3)** if exact equal replication is not essential.
   M3 does not require the slot identity and tolerates odd $J^*$ freely.
4. **Use `allow_approximate = TRUE`** as a fallback — the allocation proceeds
   with the closest possible balance, accepting minor replication differences.

```{r slot-identity-check, eval=FALSE}
# Quick check: is your combination feasible?
# J* = 75 (odd), I = 4, r = 2 -- should give --
check_balanced_incomplete_feasibility(
  n_treatments_total             = 83,   # J = J* + C = 75 + 8
  n_environments                 = 4,
  n_test_entries_per_environment = 30,   # k* guess: 30 - 8 = 22, 4*22=88 != 75*2=150
  target_replications            = 2,
  n_common_treatments            = 8
)
# feasible = FALSE -> adjust

# Fix: change C from 8 to 9 -> J* = 74 (even), r=2: 74*2/4 = 37
check_balanced_incomplete_feasibility(
  n_treatments_total             = 83,
  n_environments                 = 4,
  n_test_entries_per_environment = 46,   # k* = 37, k = 37+9 = 46
  target_replications            = 2,
  n_common_treatments            = 9
)
# feasible = TRUE
```


---

# 7. Genetic Structure-Aware Allocation

Standard allocation treats lines as exchangeable — any line could go to any
environment with equal probability. This is inadequate when:

- Genomic prediction is a downstream objective (related lines clustered in the
  same environments provide redundant rather than complementary information)
- Family structure is strong and environments differ in their genetic
  composition
- Genetic disconnectedness would bias G×E estimation

`OptiSparseMET` incorporates genetic structure through three mechanisms,
controlled by `allocation_group_source` in `allocate_sparse_met()`.

## 7.1 Family-based allocation (`"Family"`)

Family labels (e.g., full-sib or half-sib families) are read from
`treatment_info$Family`. Lines from the same family are treated as a group and
the allocator distributes groups across environments rather than concentrating
them. This prevents any one environment from containing a disproportionate
fraction of one family's members, which would inflate within-family variance
estimates and reduce the effective diversity of each site's test set.

## 7.2 GRM-based allocation (`"GRM"`)

When marker data are available, the genomic relationship matrix (GRM) encodes
pairwise genetic similarity between all candidate lines. The allocator performs
principal component analysis (PCA) on the GRM, then clusters lines by their positions in genomic principal
component space using k-means or hierarchical clustering. The resulting cluster
labels guide allocation just as family labels do.

This approach is more precise than family-based allocation when families are
heterogeneous in size or when marker data captures structure not visible from
pedigree labels alone. It directly improves the conditions under which genomic
best linear unbiased prediction (GBLUP)
models operate by ensuring that genomically similar lines are spread across
environments.

## 7.3 Pedigree-based allocation (`"A"`)

When genomic data are unavailable, the pedigree numerator relationship matrix
$A$ plays the same role as the GRM. PCA is performed on $A$ and clustering
proceeds identically. The resulting groups reflect pedigree relatedness rather
than marker-estimated genomic similarity.

## 7.4 Why this matters for prediction

Under a GBLUP model, prediction accuracy for unobserved lines depends on
their relatedness to observed lines. When genetically similar lines are
clustered in the same environments, the model gets redundant replication
of the same genetic profiles while other profiles go unobserved. Spreading
related lines across environments ensures the full genetic diversity of the
candidate set is represented at each site, maximizing the amount of novel
genetic information available for calibrating the prediction model.

```{r grm-allocation, eval=FALSE}
alloc_grm <- allocate_sparse_met(
  treatments                     = x$treatments,
  environments                   = x$environments,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = 39,
  target_replications            = 1,
  common_treatments              = x$common_treatments,
  allocation_group_source        = "GRM",
  GRM                            = x$OptiSparseMET_GRM,
  treatment_info                 = x$treatment_info,
  group_method                   = "kmeans",
  group_seed                     = 42,
  group_attempts                 = 25,
  min_groups_per_environment     = 4,
  min_env_per_group              = 2,
  balance_groups_across_env      = TRUE,
  force_group_connectivity       = TRUE,
  seed                           = 123
)

alloc_grm$group_by_environment  # how many lines per group per environment
alloc_grm$group_overlap_matrix  # how many groups are shared across env pairs
```

## 7.5 From environmental observations to `Sigma_E`

An environmental relationship is a description of sites; it is not, by
itself, an estimate of genetic covariance. The recommended workflow therefore
keeps the sources separate:

```{r, eval=FALSE}
dedicated_interaction_hypotheses <- data.frame(
  interaction = c(
    "temperature_x_density", "temperature_x_density"
  ),
  parent = c("temperature", "density"),
  modality = c("weather", "soil"),
  variable = c("T2M_mean_reproductive", "bulk_density_root")
)

kb <- build_environment_kernels(
  weather = historical_weather_features,
  soil = root_zone_soil_features,
  management = planned_management,
  geography = site_coordinates,
  bandwidth_multipliers = c(0.5, 1, 2),
  include_interactions = TRUE,
  variable_interactions = dedicated_interaction_hypotheses
)

variable_evidence <- assess_variable_interactions(
  kb$kernels,
  historical = adjusted_historical_MET_values,
  year_col = "year",
  p_adjust = "holm",
  n_boot = 500,
  seed = 2026
)

ec <- calibrate_environment_covariance(
  kb$kernels,
  historical = adjusted_historical_MET_values,
  year_col = "year",
  eligible_interactions = variable_evidence$supported,
  ridge = 0
)
Sigma_E <- ec$Sigma_E
ec$status
ec$cross_validation
```

`include_interactions = TRUE` constructs all available pairwise
weather-by-soil, weather-by-management, and soil-by-management kernels. The
default is a centred functional analysis of variance (functional-ANOVA)
construction,
\((H K_a H)\circ(H K_b H)\), which represents joint deviations after removing
the constant feature-space baseline from each parent. This is preferable to
the legacy raw product because it reduces main-effect leakage. Explicit
higher-order hypotheses can be supplied with `interaction_terms`, while
`interaction_mode = "product"` remains available for reproducibility with
older analyses. These are candidate covariance structures—not evidence that a
biological interaction exists. Historical MET responses must support their
contribution through a paired blocked-validation comparison against a
main-effects-only model. Unsupported interaction weights are fixed at zero.
Every interaction kernel remains a separate structural uncertainty scenario,
including when its central weight is zero or cannot be estimated. Without
historical responses, interactions never enter the central covariance, but
they are evaluated without weights through maximin robust design. This avoids
equating absence of response evidence with absence of a physical interaction.
The production workflow propagates that same uncertainty set through capacity
selection, common-treatment optimisation, allocation refinement, and final
robust evaluation. Across-year weather analysis also retains the most atypical
observed-year weather-by-soil and weather-by-management structures.

The package distinguishes omnibus modality interactions from dedicated
variable interactions. A Gaussian weather kernel represents nonlinear
functions and interactions among all weather variables implicitly. Similarly,
`weather_x_soil` contains the full weather-by-soil tensor-product hypothesis.
Neither object isolates a named combination. Therefore,
`variable_interactions` should be used when a hypothesis such as reproductive
temperature by root-zone bulk density must be guaranteed and audited
separately.

The recommended input is a long data frame:

| Column | Type | Required | Meaning |
|---|---|---:|---|
| `interaction` | character | yes | Unique hypothesis label |
| `parent` | character | yes | Parent label within that hypothesis |
| `modality` | character | yes | Covariate-block name |
| `variable` | character | yes | Exact post-quality-control variable name |

For example:

| interaction | parent | modality | variable |
|---|---|---|---|
| temperature_x_density | temperature | weather | T2M_mean_reproductive |
| temperature_x_density | density | soil | bulk_density_root |
| rainfall_x_clay | rainfall | weather | PRECTOTCORR_sum_reproductive |
| rainfall_x_clay | clay | soil | clay_root |

Each named interaction is constructed from the row-wise tensor product of its
standardised parent variables. By default, the tensor features are
residualised against the intercept and all parent main effects before the
Gaussian bandwidth-ensemble kernel is built. Strong heredity retains the exact
parent-variable kernels. The returned `variable_interaction_audit`,
`variable_ledger`, and `kernel_hierarchy` document the matched variables,
tensor dimensions, residual information, effective rank, realised bandwidths,
and parent-child relationships.

`assess_variable_interactions()` compares each dedicated interaction with the
same main-effect model. Leave-year-out validation is preferred when
year-specific historical targets exist; otherwise leave-environment-out
validation is used. Holm correction controls multiplicity. Positive mean
prediction improvement, an adjusted paired sign test below the declared
threshold, and a positive covariance contribution are all required before an
interaction becomes eligible for the central covariance. Bootstrap inclusion
quantifies stability. Screened interactions remain separate structural
uncertainty scenarios.

If historical adjusted genetic values are unavailable, no interaction is
labelled supported or unsupported. `assess_variable_interactions()` returns
`decision = "structural_uncertainty"`, while
`calibrate_environment_covariance()` retains identity as the central
covariance and returns every broad and dedicated kernel as an unweighted
maximin candidate. Hence, a physical hypothesis is represented without being
converted into arbitrary genetic covariance.

Each source passes through `qc_environmental_data()`, which returns explicit
coverage, range, provenance, and imputation ledgers. The separate-kernel
construction means a 150-column weather block does not receive fifteen times
the weight of a ten-column soil block. Gaussian bandwidth sensitivity is
integrated within each modality. Modalities are combined as genetic covariance
only when historical MET values support estimation of their weights.

For rice, `rice_weather_features()` supports actual phenological stages or
thermal time and derives vapour-pressure deficit (VPD),
precipitation-minus-evapotranspiration (P-ET) balance, hot nights,
heat/cold severity, root
water deficit, and consecutive stress spells. `fetch_soilgrids()` can retrieve
several depths and Q0.05/Q0.5/Q0.95 predictions;
`soil_profile_features()` produces thickness-weighted root-zone summaries,
uncertainty widths, available water, neighbourhood heterogeneity, and
compositional texture coordinates.

If historical adjusted genetic values are unavailable,
`calibrate_environment_covariance()` returns `status = "no_historical_met"`,
`weights = NULL`, and identity as the central `Sigma_E`. Every modality kernel
and every interaction kernel is retained separately as a sensitivity structure for
`robust_scenarios(Sigma_E_candidates = ...)`; no arbitrary mixture is formed.
For descriptive mega-environment clustering,
`consensus_environment_kernels()` combines within-kernel ranks by their median,
without elicited weights. This descriptive consensus must not be called genetic
covariance. Use maximin aggregation across the separate structures when no MET
evidence is available, rather than assigning them arbitrary scenario
probabilities. The same maximin rule is retained when historical calibration
exists because rejected or weakly identified interaction structures still lack
defensible probabilities. Across-year weather consensus is computed within the weather
modality only, so instability in weather never removes soil, management, or
geography.

The number of descriptive groups should normally be estimated with
`infer_mega_environments()`, not supplied beforehand. It compares hierarchical
clustering and k-means on kernel principal coordinates over every feasible
group count. K-means uses finite objective descent from multiple non-empty
starts, guaranteeing termination without an iteration-limit failure. Minimum
group size and silhouette separation screen out weak solutions; adjusted-Rand agreement across methods, modalities, years, and
relationship-level bootstrap consensuses screens out unstable solutions.
Years can be assigned to a common evidence block with
`relationship_groups`, preventing a long weather archive from acting as an
implicit weather weight. No weighted index chooses the winning number of
groups. If none passes all requirements, the result is an explicitly
unpartitioned network. With only one evidence block any accepted result is
labelled provisional, and hard genetic mega-environment claims still require
historical MET response evidence.

---

# 8. Common Treatments

Common treatments are lines assigned to **every** environment before sparse
allocation begins. If $C$ is the number of common treatments, each environment
has its effective sparse capacity reduced:

$$k_e^* = k_e - C$$

and the number of sparse-allocatable lines is:

$$J^* = J - C$$

## 8.1 Why common treatments matter

Common treatments serve two statistically distinct purposes.

**Design-based connectivity**: Two environments that share no tested lines can
only be linked through the model — through the assumed structure of $K$.
Common treatments provide direct comparisons that do not depend on any model
assumption. This is important when environments are weakly genetically correlated
or when the researcher wants robust cross-site inference.

**Estimation stability**: Common treatments appear in every environment, so
their effects are estimated with the highest precision. They function as
stable benchmarks against which the performance of sparse treatments is
measured. In practice, they are often elite lines or standard checks whose
behaviour is already partially known.

> **For novice readers**: Common treatments are like reference points in a
> measurement system. Even if two environments test completely different sets
> of experimental lines, they can still be compared through their shared
> reference lines — just as two labs can compare results if they all measure
> the same control sample.

## 8.2 What is selected during planning and fixed after selection

The package does not impose one universal common-set size. This does not mean
that common-genotype identities remain changeable after a design is released.
Two planning routes are available:

1. If biological, operational, or external evidence has already identified the
   anchors, supply those identities through `common_treatments`. Allocation
   treats the supplied set as fixed and places every member in every
   environment.
2. If the set has not been chosen, `suggest_common_treatments()` provides a
   feasible starting point and `optimize_common_treatments()` compares nested
   sizes and candidate identities. The optimizer chooses a larger set only when
   its gain in robust information justifies the loss of sparse testing breadth.

Once an optimized result is accepted and passed to allocation, its membership
is fixed in exactly the same way as a user-supplied set. Re-running the planning
analysis with changed evidence, capacities, seed inventory, or uncertainty
scenarios can produce a different proposal; the released fieldbook does not
adapt or change by itself.

More common treatments strengthen direct connectivity and can improve
precision, but they also consume plots and seed that would otherwise increase
testing breadth. The planning decision is therefore made jointly with genotype
identity. Local plot replication is deliberately decided later.

## 8.3 Robust common-set selection and separate local replication

Presence and replication are different decisions. A treatment is *common* when
its presence is fixed at every environment in the selected design; it may have
one plot at one site, two at another, and more at a third when the local design
and seed inventory permit.
For production planning, `optimize_common_treatments()` compares nested common
sets across residual-variance and `Sigma_E` uncertainty scenarios. Its
criterion combines robust TPE reliability, pairwise environment connectivity,
genetic effective sample size, and retained testing breadth. Family/group
coverage, full-population coverage, entry capacities, and the network-wide seed
needed for one presence plot at every environment are hard constraints.

Pairwise correlation differences affect the effective shared-treatment target
for every environment pair. `pair_aggregate = "maximin"` is the default and
protects the least-connected pair. `"cvar"` protects the weakest lower tail,
whereas `"mean"` maximises average pairwise attainment. These are pairwise
scoring rules for one global common set. Repeating a genotype at one or both
sites does not increase the shared-treatment count. Genetic effective sample
size evaluates diversity separately; it does not replace the literal number of
distinct shared identities.

```{r, eval=FALSE}
common_opt <- optimize_common_treatments(
  G = G, Sigma_E = Sigma_E,
  treatment_info = treatment_info,
  seed_info = seed_info,
  # Multiply per-plot seed by any mandatory local plot count first.
  seed_required_per_plot = mandatory_seed_per_presence,
  entry_capacities = site_entries,
  minimum_seed_buffer = 0,
  aggregate = "cvar",
  pair_aggregate = "maximin"
)

common_opt$common_presence
common_opt$pareto
```

`common_presence` is binary and records the chosen global core. Pass
`common_opt$selected` to allocation. Local one-, two-, or higher-plot
replication is then assigned by `plan_sparse_met_design()` or
`assign_replication_by_seed()` from the seed remaining after network
allocation.

Connectivity itself is evaluated from distinct shared genotypes in the binary
allocation matrix. A genotype contributes one shared treatment to an
environment pair whether it has one plot or several at either site. Supplying
`Sigma_E` to `allocate_sparse_met()` or `plan_sparse_met_design()` activates
correlation-adaptive, margin-preserving swaps: weakly connected pairs can
receive more distinct shared sparse treatments, whereas strongly connected
pairs may require fewer. Plot replication remains a downstream seed-feasible
precision decision.

---

# 9. Seed-Aware Replication

## 9.1 The feasibility constraint

A design that requests replication without accounting for available seed cannot
be deployed. Across a network, the same inventory must not be reset at every
site. A genotype consumes zero seed at any site where it is absent; where it is
present, its consumption is accumulated across all delivered plots. The joint
feasibility condition for genotype $i$ is:

$$s_i \geq \sum_e r_{ie}q_e + b$$

where:
- $s_i$ = seed available for genotype $i$, in grams
- $r_{ie}$ = delivered plots in environment $e$
- $q_e$ = grams required per plot in environment $e$
- $b$ = optional reserve in grams

The default reserve is 0 g. A positive value should be supplied only when the
breeding programme has an explicit retention, germination, or replanting
policy; it is not imposed as a statistical requirement.

## 9.2 Replication roles

`plan_sparse_met_design()` enforces the joint condition during allocation and
local replication, then reconstructs a final `seed_ledger` from all fieldbooks.
For a single local environment, `assign_replication_by_seed()` evaluates the
corresponding one-site condition and assigns one of three roles:

- **`"p_rep"`** — sufficient seed for $r_i$ plots; assigned
  `desired_replications` plots
- **`"unreplicated"`** — sufficient seed for one plot but not for
  `desired_replications` plots (downgraded)
- **`"excluded"`** — insufficient seed even for one plot

Three design modes control how these roles are assigned:

| Mode | `"augmented"` | `"p_rep"` | `"rcbd_type"` |
|------|--------------|-----------|---------------|
| Target | 1 plot per line | $r$ plots for a feasible subset | $r$ plots for all lines |
| Downgrade | Excluded if seed < $q_e$ | Non-candidates get 1 plot | Yes, if `shortage_action = "downgrade"` |
| P-rep selection | — | By priority (seed, order, random) | — |

```{r seed-replication, eval=FALSE}
seed_df <- data.frame(
  Treatment     = x$treatments,
  SeedAvailable = sample(20:120, length(x$treatments), replace = TRUE)
)

rep_plan <- assign_replication_by_seed(
  treatments             = x$treatments,
  seed_available         = seed_df,
  seed_required_per_plot = 10,
  replication_mode       = "p_rep",
  desired_replications   = 2,
  priority               = "seed_available",
  max_prep               = 20,
  minimum_seed_buffer    = 5,
  shortage_action        = "downgrade",
  seed                   = 123
)

length(rep_plan$p_rep_treatments)        # replicated at 2 reps
length(rep_plan$unreplicated_treatments) # assigned 1 plot (downgraded)
length(rep_plan$excluded_treatments)     # insufficient seed
rep_plan$seed_summary                    # full per-line feasibility table
```

---

# 10. Within-Environment Design Engines

## 10.1 `met_prep_famoptg()` — block-based repeated-check designs

This engine constructs repeated-check block designs for environments where
the primary local control mechanism is blocking (incomplete block designs).
It supports three design classes:

**Augmented repeated-check design**

All non-check lines are unreplicated. Each block contains the same set of check
treatments plus a unique subset of experimental lines. This design is
appropriate for early screening stages when the number of candidates far exceeds
available resources and seed is very limited. The checks allow estimation of
block effects, which are subtracted to improve comparisons among the
unreplicated lines.

**Partially replicated (p-rep) design**

A subset of lines — those with more seed or higher priority — is replicated
at `desired_replications` plots, distributed across distinct blocks. The
remainder appear unreplicated. P-rep designs are efficient because they
concentrate replication where it provides the most information while still
including a broad set of candidates.

**RCBD-type repeated-check design**

All non-check lines are replicated at a uniform level. This is the closest to
a standard RCBD within a single environment, extended to include repeated
checks in every block.

**Structural guarantees enforced by `met_prep_famoptg()`**:

- Check treatments appear in every block
- Replicated lines appear in distinct blocks (at most once per block)
- Unreplicated lines appear exactly once in the design

Optional features include family-based or GRM/A-based genetic dispersion
optimisation (minimising the average relatedness of adjacent plots) and
mixed-model efficiency evaluation (A, D, CDmean).

```{r prep-famoptg, eval=FALSE}
design_E1 <- met_prep_famoptg(
  check_treatments        = c("CHK1", "CHK2"),
  check_families          = c("CHECK", "CHECK"),
  p_rep_treatments        = rep_plan$p_rep_treatments,
  p_rep_reps              = rep_plan$p_rep_reps,
  p_rep_families          = x$treatment_info$Family[
                              match(rep_plan$p_rep_treatments,
                                    x$treatment_info$Treatment)],
  unreplicated_treatments = rep_plan$unreplicated_treatments,
  unreplicated_families   = x$treatment_info$Family[
                              match(rep_plan$unreplicated_treatments,
                                    x$treatment_info$Treatment)],
  n_blocks     = 4L,
  n_rows       = 10L,
  n_cols       = 12L,
  order        = "row",
  serpentine   = TRUE,
  seed         = 123
)

design_E1$layout_matrix  # plot-level field map
design_E1$field_book     # row-level data frame ready for export
```

## 10.2 `met_alpha_rc_stream()` — row-column alpha designs

This engine constructs alpha row-column designs for large fields where both
row and column gradients are present. It is most appropriate when:

- The field is large and rectangular
- Row and column variation is substantial relative to treatment differences
- A first-order autoregressive spatial residual model in one dimension (AR1) or
  two dimensions (AR1×AR1) will be used at analysis

The design is built by streaming plots across the field in a specified
traversal order (row-major, column-major, or serpentine). Each replicate
occupies a contiguous block of rows or columns. Within each replicate, entries
are partitioned into incomplete blocks, with check treatments inserted at
regular intervals within every incomplete block.

**Structural guarantees**:

- Each entry appears exactly once per replicate
- Check treatments appear in every incomplete block
- Block sizes are constrained within `[min_block_size, max_block_size]`

```{r alpha-rc, eval=FALSE}
design_E2 <- met_alpha_rc_stream(
  check_treatments = c("CHK1", "CHK2"),
  check_families   = c("CHECK", "CHECK"),
  entry_treatments = rep_plan$p_rep_treatments,
  entry_families   = x$treatment_info$Family[
                       match(rep_plan$p_rep_treatments,
                             x$treatment_info$Treatment)],
  n_reps         = 2L,
  n_rows         = 10L,
  n_cols         = 12L,
  min_block_size = 8L,
  max_block_size = 12L,
  order          = "row",
  serpentine     = TRUE,
  seed           = 123
)

design_E2$layout_matrix
design_E2$field_book
design_E2$design_info   # n_reps, n_blocks_per_rep, total_used_plots
```

## 10.3 Design engine selection guide

Use `met_prep_famoptg()` when:

- Repeated check distribution across blocks is a structural requirement
- The design needs to support p-rep, augmented, or RCBD-type structure
- The field is subdivided into blocks but row-column spatial modelling is
  not the primary concern

Use `met_alpha_rc_stream()` when:

- The field is large and continuous, with both row and column gradients
- Spatial models (AR1 or AR1×AR1) will be used at analysis
- The design needs fixed-grid compatibility for automated plot management

The engines can be mixed across environments within the same trial.

---

# 11. Efficiency Evaluation

## 11.1 Block-based designs: `met_evaluate_famoptg_efficiency()`

For designs produced by `met_prep_famoptg()`, efficiency is evaluated under a
model with a block random effect:

$$y = X\beta + Zg + Wb + e$$

where $b \sim N(0,\, I\sigma_b^2)$ is the block effect. The variance
components are specified in `varcomp` using `sigma_b2` for the block variance.
Note that this model has a **single block variance component** — there is no
separate replicate or incomplete-block-within-replicate structure.

The function computes A-criterion, D-criterion (for fixed treatment effects),
and CDmean (for random treatment effects), under any of three residual
structures: `"IID"`, `"AR1"` (along rows or columns), or `"AR1xAR1"` (spatial
gradient in both dimensions).

```{r eval-famoptg, eval=FALSE}
eff_E1 <- met_evaluate_famoptg_efficiency(
  field_book       = design_E1$field_book,
  n_rows           = 10L,
  n_cols           = 12L,
  check_treatments = c("CHK1", "CHK2"),
  varcomp = list(
    sigma_e2  = 1.0,   # residual variance
    sigma_g2  = 0.5,   # genetic variance (for CDmean)
    sigma_b2  = 0.3,   # block variance
    sigma_r2  = 0.05,  # row variance (for spatial structures)
    sigma_c2  = 0.05   # column variance
  ),
  treatment_effect    = "random",
  prediction_type     = "IID",
  residual_structure  = "IID"
)

eff_E1$CDmean     # in [0, 1]: higher is better
eff_E1$mean_PEV   # mean prediction error variance
eff_E1$A          # A-efficiency (1 / A_criterion): higher is better
eff_E1$D          # D-efficiency: higher is better
```

## 11.2 Row-column designs: `met_evaluate_alpha_efficiency()`

For designs produced by `met_alpha_rc_stream()`, efficiency is evaluated under
a model with replicate and incomplete-block-within-replicate random effects:

$$y = X\beta + Zg + W_r r + W_{b} b + e$$

where $r \sim N(0,\, I\sigma_{\text{rep}}^2)$ and
$b \sim N(0,\, I\sigma_{\text{ib}}^2)$ are the replicate and incomplete block
effects respectively. The `varcomp` list must contain `sigma_rep2` and
`sigma_ib2` (not `sigma_b2`). This two-level blocking structure reflects the
hierarchical design of alpha row-column layouts.

```{r eval-alpha, eval=FALSE}
eff_E2 <- met_evaluate_alpha_efficiency(
  field_book       = design_E2$field_book,
  n_rows           = 10L,
  n_cols           = 12L,
  check_treatments = c("CHK1", "CHK2"),
  varcomp = list(
    sigma_e2   = 1.0,
    sigma_g2   = 0.5,
    sigma_rep2 = 0.4,   # replicate variance
    sigma_ib2  = 0.2,   # incomplete block within replicate
    sigma_r2   = 0.05,
    sigma_c2   = 0.05
  ),
  treatment_effect   = "random",
  prediction_type    = "IID",
  residual_structure = "IID"
)

eff_E2$CDmean
eff_E2$mean_PEV
```

> **Key distinction**: `met_evaluate_famoptg_efficiency()` uses `sigma_b2`
> (one block level); `met_evaluate_alpha_efficiency()` uses `sigma_rep2` +
> `sigma_ib2` (two block levels: replicate and incomplete block within
> replicate). Using the wrong function for the wrong design will produce an
> error or incorrect results.

---

# 12. Design Optimisation

Once a base design is constructed and its efficiency evaluated, the optimizer
functions search for a better arrangement of the same treatments within the
same blocking structure.

## 12.1 `met_optimize_famoptg()` — block-based optimisation

`met_optimize_famoptg()` uses Random Restart (RS): it generates multiple
independent random designs, evaluates each, and returns the best. All
structural guarantees of `met_prep_famoptg()` are enforced in every candidate
design.

```{r optimize-famoptg, eval=FALSE}
opt_E1 <- met_optimize_famoptg(
  # All arguments of met_prep_famoptg(), plus:
  check_treatments        = c("CHK1", "CHK2"),
  check_families          = c("CHECK", "CHECK"),
  p_rep_treatments        = rep_plan$p_rep_treatments,
  p_rep_reps              = rep_plan$p_rep_reps,
  p_rep_families          = x$treatment_info$Family[
                              match(rep_plan$p_rep_treatments,
                                    x$treatment_info$Treatment)],
  unreplicated_treatments = rep_plan$unreplicated_treatments,
  unreplicated_families   = x$treatment_info$Family[
                              match(rep_plan$unreplicated_treatments,
                                    x$treatment_info$Treatment)],
  n_blocks            = 4L,
  n_rows              = 10L,
  n_cols              = 12L,
  # Optimisation arguments:
  treatment_effect    = "random",
  prediction_type     = "IID",
  criterion           = "CDmean",
  varcomp = list(sigma_e2=1, sigma_g2=0.5, sigma_b2=0.3,
                 sigma_r2=0.05, sigma_c2=0.05),
  n_restarts          = 20L,
  seed                = 123
)

opt_E1$optimization$best_score    # best CDmean found
opt_E1$optimization$score_history # CDmean at each restart
opt_E1$optimization$n_failed      # restarts that produced invalid designs
opt_E1$field_book                 # the best design's field book
opt_E1$efficiency                 # efficiency metrics of the best design
```

## 12.2 `met_optimize_alpha_rc()` — row-column optimisation

`met_optimize_alpha_rc()` supports three optimisation algorithms:

**RS (Random Restart)**: Same as `met_optimize_famoptg()`. Generates multiple
independent candidates and keeps the best. Simple and parallelizable.

**SA (Simulated Annealing)**: Starts from one design and iteratively perturbs
it. At each step, a worse design is accepted with probability
$\exp(-\Delta / T)$, where $\Delta$ is the score decrease and $T$ is the
current temperature. As $T$ decreases from `sa_temp_start` to `sa_temp_end`,
the algorithm transitions from broad exploration to local refinement. SA is
effective when the fitness landscape has many local optima.

**GA (Genetic Algorithm)**: Maintains a population of designs. At each
generation, pairs of high-scoring designs are combined ("crossover") and
random perturbations are applied ("mutation"). Elite designs are carried
forward unchanged. GA explores a broader search space than RS or SA for the
same number of objective function evaluations, at the cost of higher
implementation complexity.

```{r optimize-alpha, eval=FALSE}
# Simulated Annealing
opt_E2_sa <- met_optimize_alpha_rc(
  check_treatments = c("CHK1", "CHK2"),
  check_families   = c("CHECK", "CHECK"),
  entry_treatments = rep_plan$p_rep_treatments,
  entry_families   = x$treatment_info$Family[
                       match(rep_plan$p_rep_treatments,
                             x$treatment_info$Treatment)],
  n_reps            = 2L,
  n_rows            = 10L,
  n_cols            = 12L,
  method            = "SA",
  treatment_effect  = "random",
  prediction_type   = "IID",
  criterion         = "CDmean",
  varcomp = list(sigma_e2=1, sigma_g2=0.5, sigma_rep2=0.4,
                 sigma_ib2=0.2, sigma_r2=0.05, sigma_c2=0.05),
  sa_max_iter       = 500L,
  sa_temp_start     = 1.0,
  sa_temp_end       = 0.01,
  seed              = 123
)

opt_E2_sa$optimization$best_score
opt_E2_sa$optimization$sa_params   # SA-specific settings
opt_E2_sa$optimization$ga_params   # NULL (not used)

# Genetic Algorithm
opt_E2_ga <- met_optimize_alpha_rc(
  check_treatments = c("CHK1", "CHK2"),
  check_families   = c("CHECK", "CHECK"),
  entry_treatments = rep_plan$p_rep_treatments,
  entry_families   = x$treatment_info$Family[
                       match(rep_plan$p_rep_treatments,
                             x$treatment_info$Treatment)],
  n_reps            = 2L,
  n_rows            = 10L,
  n_cols            = 12L,
  method            = "GA",
  treatment_effect  = "random",
  prediction_type   = "IID",
  criterion         = "CDmean",
  varcomp = list(sigma_e2=1, sigma_g2=0.5, sigma_rep2=0.4,
                 sigma_ib2=0.2, sigma_r2=0.05, sigma_c2=0.05),
  ga_pop_size       = 20L,
  ga_n_generations  = 50L,
  ga_elitism        = 4L,
  seed              = 123
)

opt_E2_ga$optimization$best_score
opt_E2_ga$optimization$ga_params
```

---

# 13. Full Pipeline Example

## 13.1 Step-by-step

```{r full-pipeline-steps, eval=FALSE}
library(OptiSparseMET)

data("OptiSparseMET_example_data")
x <- OptiSparseMET_example_data

# Stage 0: verify capacity
k_safe <- suggest_safe_k(x$treatments, x$environments,
                         common_treatments = x$common_treatments,
                         buffer = 3)   # 39

# Stage 1: allocation
# k_safe = 39 does not satisfy the slot identity for M4 (112*1=112 ≠ 4*31=124).
# Use random_balanced (M3) here, or adjust k for M4 slot identity.
# For M4 with r=1: k* = 112/4 = 28, k = 36.
alloc <- allocate_sparse_met(
  treatments                     = x$treatments,
  environments                   = x$environments,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = k_safe,
  target_replications            = 1,
  common_treatments              = x$common_treatments,
  allocation_group_source        = "Family",
  treatment_info                 = x$treatment_info,
  seed                           = 123
)

alloc$summary
alloc$overlap_matrix   # environments x environments: how many lines shared

# Stage 2: seed-aware replication
rep_plan <- assign_replication_by_seed(
  treatments             = x$treatments,
  seed_available         = x$seed_info,
  seed_required_per_plot = 8,
  replication_mode       = "p_rep",
  desired_replications   = 2,
  max_prep               = 15,
  shortage_action        = "downgrade",
  seed                   = 123
)

# Stage 3: design per environment (example for Env_A)
design_A <- met_prep_famoptg(
  check_treatments        = x$env_design_specs$Env_A$check_treatments,
  check_families          = x$env_design_specs$Env_A$check_families,
  p_rep_treatments        = rep_plan$p_rep_treatments,
  p_rep_reps              = rep_plan$p_rep_reps,
  p_rep_families          = x$treatment_info$Family[
                              match(rep_plan$p_rep_treatments,
                                    x$treatment_info$Treatment)],
  unreplicated_treatments = rep_plan$unreplicated_treatments,
  unreplicated_families   = x$treatment_info$Family[
                              match(rep_plan$unreplicated_treatments,
                                    x$treatment_info$Treatment)],
  n_blocks   = x$env_design_specs$Env_A$n_blocks,
  n_rows     = x$env_design_specs$Env_A$n_rows,
  n_cols     = x$env_design_specs$Env_A$n_cols,
  seed       = 123
)

# Stage 4: combine field books
met_book <- combine_met_fieldbooks(
  field_books       = list(Env_A = design_A$field_book),
  local_designs     = c(Env_A = "met_prep_famoptg"),
  replication_modes = c(Env_A = "p_rep"),
  sparse_method     = "equireplicate",
  common_treatments = x$common_treatments
)

head(met_book)
```

## 13.2 End-to-end pipeline

`plan_sparse_met_design()` executes all four stages in a single call. Each
environment's local design is controlled by `env_design_specs`, a named list
where the `design` field specifies `"met_prep_famoptg"` or
`"met_alpha_rc_stream"`:

```{r pipeline-full, eval=FALSE}
out <- plan_sparse_met_design(
  treatments                     = x$treatments,
  environments                   = x$environments,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = 39,
  target_replications            = 1,
  common_treatments              = x$common_treatments,
  env_design_specs               = x$env_design_specs,
  treatment_info                 = x$treatment_info,
  seed_info                      = x$seed_info,
  seed_required_per_plot         = x$seed_required_per_plot,
  minimum_seed_buffer            = 1,
  seed                           = 123
)

names(out)
# "sparse_allocation"    "environment_designs"   "combined_field_book"
# "environment_summary"  "group_environment_summary"
# "efficiency_summary"   "seed_ledger"  "summary"  "seed_used"

head(out$combined_field_book)
out$environment_summary         # one row per environment
out$efficiency_summary          # efficiency metrics per environment
out$seed_ledger                 # grams consumed across all final fieldbooks
```

## 13.3 Environment-specific design specifications

The `env_design_specs` list allows each environment to use a different design
engine and replication mode. The `design` field routes to the correct
constructor. Additional fields are passed directly as arguments:

```{r env-specs, eval=FALSE}
env_design_specs <- list(

  # p-rep block design
  Env_A = list(
    design               = "met_prep_famoptg",
    replication_mode     = "p_rep",
    desired_replications = 2L,
    max_prep             = 15L,
    shortage_action      = "downgrade",
    check_treatments     = c("CHK1", "CHK2"),
    check_families       = c("CHECK", "CHECK"),
    n_blocks             = 4L,
    n_rows               = 10L,
    n_cols               = 12L,
    cluster_source       = "Family"
  ),

  # Row-column alpha design
  Env_B = list(
    design                    = "met_alpha_rc_stream",
    check_treatments          = c("CHK1", "CHK2"),
    check_families            = c("CHECK", "CHECK"),
    n_reps                    = 2L,
    n_rows                    = 10L,
    n_cols                    = 12L,
    min_block_size            = 10L,
    cluster_source            = "Family"
  ),

  # Augmented repeated-check design
  Env_C = list(
    design           = "met_prep_famoptg",
    replication_mode = "augmented",
    check_treatments = c("CHK1", "CHK2"),
    check_families   = c("CHECK", "CHECK"),
    n_blocks         = 4L,
    n_rows           = 10L,
    n_cols           = 10L,
    cluster_source   = "Family"
  )
)
```

---

# 14. Connectivity and Genomic Prediction: Simulation Illustration

Cross-environment connectivity only matters when genotype performance varies
across environments -- that is, under genotype-by-environment interaction
(G x E). We therefore evaluate designs under a genuine G x E model in which the
genotype-by-environment effects have covariance `Sigma_E %x% G` (a genetic
correlation structure across environments, not a single shared breeding value).
`met_information()` returns the coupled across-TPE prediction-error variance
(PEV) and reliability (CDmean); `simulate_met()` returns the realised selection
accuracy and genetic gain those designs achieve, together with Monte Carlo
standard errors and 95% intervals. When fieldbooks exist, pass their actual
treatment-by-environment plot counts through `reps`; do not substitute an
intended average replication.

```{r simulation, eval=FALSE}
set.seed(1)
n_geno <- 120
n_env  <- 4
lines  <- paste0("L", seq_len(n_geno))
envs   <- paste0("E", seq_len(n_env))

# Genomic relationship matrix
M  <- matrix(rnorm(n_geno^2), n_geno)
G0 <- tcrossprod(M) / n_geno
diag(G0) <- diag(G0) + 0.5
dimnames(G0) <- list(lines, lines)

# Broad TPE: environments with a mix of positive and negative genetic
# correlations (this is what makes connectivity matter).
Sigma_E <- matrix(c( 1.0,  0.6,  0.2, -0.1,
                     0.6,  1.0,  0.3,  0.0,
                     0.2,  0.3,  1.0,  0.5,
                    -0.1,  0.0,  0.5,  1.0), 4, 4)
dimnames(Sigma_E) <- list(envs, envs)

# Fully balanced positive control: every line in every environment.
alloc_full <- matrix(1L, n_geno, n_env, dimnames = list(lines, envs))

# Random sparse allocation, each line in ~2 of 4 environments.
alloc_sparse <- allocate_sparse_met(
  treatments                     = lines,
  environments                   = envs,
  allocation_method              = "random_balanced",
  n_test_entries_per_environment = 60,
  target_replications            = 2,
  allow_approximate              = TRUE,
  seed                           = 1
)$allocation_matrix

# Equireplicate allocation with correlation-adaptive distinct-treatment
# connectivity. Maximin protects the weakest target-attainment pair.
alloc_eqr <- allocate_sparse_met(
  treatments                     = lines,
  environments                   = envs,
  allocation_method              = "equireplicate",
  n_test_entries_per_environment = 60,
  target_replications            = 2,
  Sigma_E                        = Sigma_E,
  pair_aggregate                 = "maximin",
  seed                           = 1
)$allocation_matrix

# Refine the equireplicate allocation directly against the coupled G x E
# objective (minimise across-TPE PEV).
alloc_opt <- optimize_allocation_gxe(
  alloc_eqr, G = G0, Sigma_E = Sigma_E, iter = 100, seed = 1
)$allocation_matrix

# Design-based coupled PEV / CDmean (lower PEV, higher CDmean is better).
sapply(list(full = alloc_full, sparse = alloc_sparse,
            equireplicate = alloc_eqr, optimised = alloc_opt),
       function(A) met_information(A, G = G0, Sigma_E = Sigma_E)$mean_PEV)

# Realised selection accuracy and genetic gain across the TPE.
sapply(list(full = alloc_full, sparse = alloc_sparse,
            equireplicate = alloc_eqr, optimised = alloc_opt),
       function(A) simulate_met(A, G = G0, Sigma_E = Sigma_E,
                                n_sim = 50, seed = 1)$accuracy_mean)
```

The fully balanced design is the positive control and information upper bound;
its realised Monte Carlo accuracy can still vary. Among the equal-budget sparse
options, the comparison tests whether environment-pair balancing and direct
`optimize_allocation_gxe()` refinement improve prediction over plain random
sparse allocation. That result must be measured rather than assumed, and its
uncertainty should be reported. The next section extends this single-scenario
illustration to the covariance, variance, operational, and site-loss
perturbations required for a release decision.

---

# 15. Benchmarking before field release

The illustration above compares designs under one covariance and one variance
setting. A release decision requires a wider validation. The public
benchmarking framework provides: (i) known-truth environmental covariance
recovery, (ii) leave-year-out and leave-environment-out validation, (iii)
full-testing, naive random, M3, and strict-M4 comparators, (iv) paired
Monte Carlo evaluation across covariance, variance, and complete site-loss
scenarios, (v) missing-environmental-covariate stress tests, and (vi)
common-set, capacity, and allocation stability.

```{r benchmark-framework, eval=FALSE}
reference <- make_met_benchmark_designs(
  treatments = rownames(G0),
  environments = colnames(Sigma_E),
  n_test_entries_per_environment = 60,
  target_replications = 2,
  seed = 2026
)

benchmark <- benchmark_met_designs(
  designs = reference,
  G = G0,
  Sigma_E = Sigma_E,
  sigma_e2 = c(0.5, 1, 2),
  environment_dropout = list(
    none = character(),
    fourth_site_lost = "E4"
  ),
  n_sim = 200,
  seed = 2026
)

benchmark$design_summary
plot_pareto_frontier(benchmark, metric = "accuracy")
```

The paired simulation exposes every design to the same realised genetic effects
and residual standard-normal draws. The Pareto graph uses one vertical endpoint,
a 0--100% cost-minimisation axis below, and decreasing real cost above. See
`vignette("OptiSparseMET-benchmarking")` for the complete input contract and
demonstrations.

---

# 16. Design Component Summary

| Component | Governing principle | Package function(s) |
|-----------|--------------------|--------------------|
| Environmental evidence | Keep modalities and interactions separate unless historical responses support calibration | `build_environment_kernels()`, `calibrate_environment_covariance()` |
| Environmental grouping | Infer and stability-screen group number; retain one network when unsupported | `infer_mega_environments()` |
| Structural uncertainty | Evaluate unweighted alternatives by maximin robustness | `robust_scenarios()` |
| Capacity verification | All treatments must be assignable at least once | `suggest_safe_k()`, `min_k_for_full_coverage()` |
| Allocation | Balance replication depth against coverage breadth | `allocate_sparse_met()` |
| Equireplication feasibility | $\sum_e k_e^* = J^*r$ and the degree sequence must be realisable | `check_equireplicate_feasibility()` |
| Genetic structure | Prevent clustering; improve prediction conditions | `derive_allocation_groups()` |
| Connectivity | Count distinct shared genotypes; combine a global common core with correlation-adaptive pairwise overlap while keeping replication separate | `suggest_common_treatments()`, `optimize_common_treatments()`, `allocate_sparse_met()` |
| Seed feasibility | One inventory must fund allocation and replication across the network | `plan_sparse_met_design()`, `allocate_sparse_met()` |
| Block design | Control local heterogeneity within environments | `met_prep_famoptg()` |
| Row-column design | Control spatial gradients in large fixed-grid fields | `met_alpha_rc_stream()` |
| Efficiency evaluation | Quantify design quality before field deployment | `met_evaluate_famoptg_efficiency()`, `met_evaluate_alpha_efficiency()` |
| Coupled MET evaluation | Use delivered plot counts to quantify PEV, CDmean, accuracy, and gain | `met_information()`, `simulate_met()` |
| Empirical benchmarking | Compare reference designs with paired truths, operational ledgers, and covariance/variance/site-loss stress tests | `make_met_benchmark_designs()`, `benchmark_met_designs()` |
| Environmental validation | Quantify covariance recovery, blocked prediction, and missing-covariate sensitivity | `benchmark_environment_models()`, `simulate_environment_model_benchmark()`, `benchmark_environment_missingness()` |
| Decision stability | Report common-set, capacity, and allocation reproducibility | `summarize_design_stability()` |
| Optimisation | Search for higher-efficiency arrangements | `met_optimize_famoptg()`, `met_optimize_alpha_rc()` |
| Assembly | Produce a unified field book for all environments | `combine_met_fieldbooks()`, `plan_sparse_met_design()` |

---

# 17. Conclusions

Sparse MET design requires more than balanced incidence. It requires coherent
decisions about environmental uncertainty, genetic connectivity, testing
breadth, local replication, field geometry, and seed. `OptiSparseMET`
integrates these decisions through four principles: (i) environmental
main-effect and interaction kernels remain explicit rather than being collapsed
by arbitrary weights; (ii) historical responses calibrate genetic covariance
when available, while maximin sensitivity analysis protects decisions when
they are not; (iii) common-set size and identity are optimised from binary
presence, while local replication is assigned separately from remaining seed;
and (iv) delivered plot counts drive information, simulation, and cost.

The resulting workflow supports both random-balanced and equireplicate
allocation, additive and optional hybrid-dominance relationships,
block-based and row-column local designs, and prediction-oriented evaluation
through prediction-error variance (PEV), mean coefficient of determination
(CDmean), selection accuracy, and genetic gain. Unsupported environmental
grouping falls back to one network, and infeasible allocation or seed demands
are reported before deployment.

For plant breeders, the practical output is an auditable trial plan: every
genotype-site presence is explicit, every gram of seed is counted once, every
local field book respects its geometry, and every design recommendation can be
examined against explicit reference designs and plausible
quantitative-genetic scenarios before planting.

---

# 17. References

Montesinos-López, O. A., Mosqueda-González, B. A., Salinas-Ruiz, J.,
Montesinos-López, A., & Crossa, J. (2023). Sparse multi-trait genomic
prediction under balanced incomplete block design. *The Plant Genome*, 16,
e20305. <https://doi.org/10.1002/tpg2.20305>

Rincent, R., Laloe, D., Nicolas, S., Altmann, T., Brunel, D., Revilla, P.,
Rodriguez, V. M., Moreno-Gonzalez, J., Melchinger, A., Bauer, E., Schoen,
C.-C., Meyer, N., Giauffret, C., Bauland, C., Jamin, P., Laborde, J.,
Monod, H., Flament, P., Charcosset, A., & Moreau, L. (2012). Maximizing the
reliability of genomic selection by optimizing the calibration set of reference
individuals: Comparison of methods in two diverse groups of maize inbreds
(*Zea mays* L.). *Genetics*, 192, 715–728.
<https://doi.org/10.1534/genetics.112.141473>
