---
title: "End-to-End Sparse MET Pipeline: from Environmental Classification to the Combined Field Book"
author: "OptiSparseMET"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
header-includes:
  - |
    <script>
    window.MathJax = { tex: { inlineMath: [['\\(','\\)']], displayMath: [['\\[','\\]']] }, svg: { fontCache: 'global' } };
    </script>
    <script defer src="https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"></script>
vignette: >
  %\VignetteIndexEntry{End-to-End Sparse MET Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
# The pipeline below is a complete, self-contained, reproducible example on
# SIMULATED data. Chunks are not evaluated at build time so the vignette builds
# quickly and without network access (weather/soil fetches hit external
# services). To run the whole thing, set eval = TRUE, or copy the code into an
# R session. Every call uses the package's real argument names and return
# fields.
knitr::opts_chunk$set(eval = FALSE, collapse = TRUE, comment = "#>",
                      fig.width = 7, fig.height = 4.5)
library(OptiSparseMET)
set.seed(2026)
```

# What this vignette builds

`OptiSparseMET` designs **sparse multi-environment trials (METs)** by optimising
two coupled levels at once:

* **Level 1 — across-environment allocation:** *which* genotypes are tested in
  *which* environments, so the whole network is genetically connected and every
  line is predictable across the target population of environments (TPE).
* **Level 2 — within-environment field layout:** *how* each site is laid out
  (blocks, checks, alpha row-column), which sets how precisely each plot is
  measured.

This tutorial runs the **entire pipeline end to end**, organised by module, on a
realistic-but-simulated rice testcross program:

1. **Module 1 — Environmental classification (enviromics).** Build environment
   kernels from weather / soil / management / geography, calibrate a data-driven
   environmental covariance $\Sigma_E$, and cluster sites into mega-environments.
2. **Module 2 — Genetic relationship matrices.** Genomic `G` (VanRaden), a
   hybrid/testcross GRM with optional dominance, and a made-invertible tune-up.
3. **Module 3 — Sparse allocation.** Feasibility, auto-suggested common
   treatments, and M3 / M4 allocation under per-site capacities.
4. **Module 4 — Seed-aware replication.** Turn a single seed inventory into a
   feasible replication plan (p-rep / augmented).
5. **Module 5 — Within-environment field design.** Block-based repeated-check
   designs and alpha row-column designs, with efficiency evaluation.
6. **Module 6 — Coupled optimisation & information.** The two-level information
   matrix, exact multi-trait index reliability, criterion + gain + cost
   objectives, exchange/SA optimisation, and the Pareto frontier.
7. **Module 7 — Simulation.** Monte-Carlo prediction accuracy and realised gain.
8. **Module 8 — Benchmarking.** Complete / random / M3 / M4 / optimised designs
   compared with confidence intervals, decision stability, and tail risk.
9. **Module 9 — Field books & field maps.** Per-site and combined field books,
   and a field map drawn from the layout.

A one-call wrapper (`plan_sparse_met_design()`) and a validation appendix close
the vignette.

# Setup: installation and loading

Install the package from GitHub (vignettes included):

```{r install, eval = FALSE}
# install.packages("remotes")
remotes::install_github("FAkohoue/OptiSparseMET",
                        build_vignettes = TRUE, dependencies = TRUE)
```

Optional packages (all in `Suggests`) unlock specific capabilities; install only
what you need:

```{r install-optional, eval = FALSE}
# Live environmental data
install.packages(c("nasapower",               # daily weather (NASA POWER)
                   "terra", "gdalUtilities",   # SoilGrids retrieval
                   "jsonlite"))
# Genomic relationship backends and matrix conditioning
install.packages(c("AGHmatrix", "ASRgenomics"))
# Multi-trait desired-gain index constructor (from GitHub)
remotes::install_github("FAkohoue/DGQGSI")
# Optional engines: exact BIBD, random-forest interaction screening,
# parallel fetching, and mixed-model variance estimation
install.packages(c("crossdes", "ranger", "future", "future.apply", "sommer"))
```

Load the package:

```{r load, eval = TRUE, message = FALSE}
library(OptiSparseMET)
set.seed(2026)
```

The simulated example below needs only the base package (its `Imports`:
**Matrix**, **pracma**, **splines**). The `Suggests` are required only for live
weather/soil fetching, the AGHmatrix / ASRgenomics backends, the DGQGSI index,
or parallel downloads — each function that needs one checks for it and degrades
gracefully if it is absent.

# The example program

```{r program}
# ---- The breeding program (simulated) -------------------------------------
n_hybrids  <- 300L                      # testcross hybrids to evaluate
n_families <- 12L
n_markers  <- 800L
hybrids    <- sprintf("H%03d", seq_len(n_hybrids))
families   <- sample(sprintf("F%02d", seq_len(n_families)), n_hybrids, TRUE)

# Seven sites across three countries. Ecuador sites are (nominally) uncapped;
# Colombia/CIAT capped at 200 plots incl. checks; Peru at 182 plots.
sites <- c("Ecu_Daule", "Ecu_Samborondon", "Ecu_Babahoyo", "Ecu_Vinces",
           "Col_Saldana", "CIAT_Palmira", "Per_Chiclayo")
```

---

# Module 1 — Environmental classification (enviromics)

**Goal:** describe *how similar the environments are* so the design can connect
them and cluster them into mega-environments. This drives $\Sigma_E$, which every
downstream optimisation consumes.

## 1.1 Site coordinates

Weather and soil retrieval need each site's coordinates.

```{r sites-coords}
sites_xy <- data.frame(
  environment = sites,
  latitude  = c(-1.86, -1.96, -1.80, -1.55,  3.93,  3.50, -6.77),
  longitude = c(-79.98,-79.72,-79.53,-79.75,-75.02,-76.36,-79.84),
  stringsAsFactors = FALSE)
```

## 1.2 Choosing weather variables — the six defaults, or *all* of them

When `pars = NULL`, the weather fetchers return **six** core variables
(temperature, max/min temperature, precipitation, radiation, humidity). You are
not limited to those: `enviromic_variable_catalog("weather")` lists every
fetchable variable (20 in total, including NASA POWER codes for wind, pressure,
dew point, soil wetness, evapotranspiration, PAR, …). Pass the full vector to
`pars` to retrieve **all** variables instead of the default subset.

```{r weather-vars}
# The catalogue of all fetchable weather variables:
enviromic_variable_catalog("weather")[, c("variable", "units", "description")]

# The six defaults (used when pars = NULL):
six_defaults    <- head(enviromic_variable_catalog("weather")$variable, 6)

# ALL available weather variables -- use this to fetch everything:
all_weather_pars <- enviromic_variable_catalog("weather")$variable
all_weather_pars                                   # 20 variables
```

## 1.3 Download ten years of historical daily weather

`historical_envirotype()` pulls **daily** NASA POWER weather for a vector of past
years over a growing window, then summarises it into a multi-year *typical*
envirotype plus *interannual variability* and *per-year* matrices (used later for
stability analysis). Set `pars = all_weather_pars` to characterise the sites on
every variable rather than the six defaults. (Requires `nasapower` and network
access, so it is not evaluated here.)

```{r fetch-hist, eval = FALSE}
this_year <- as.integer(format(Sys.Date(), "%Y"))

hist_wx <- historical_envirotype(
  sites   = sites_xy,
  years   = (this_year - 10):(this_year - 1),   # <-- the last ten years
  window  = c("11-01", "02-28"),                 # growing season (Nov -> Feb)
  pars    = all_weather_pars,                     # <-- ALL variables (omit for the 6 defaults)
  variability = "sd",
  cache_dir = "power_cache", workers = 4)         # cache + parallel download

hist_wx$typical      # environment x feature: multi-year typical envirotype
hist_wx$variability  # interannual variability (sd) per feature
hist_wx$per_year     # list of per-year environment x feature matrices
```

For a **single** season instead of a multi-year climatology, fetch the daily
series directly and derive features. `pars = all_weather_pars` again requests
every variable:

```{r fetch-series, eval = FALSE}
sites_win <- transform(sites_xy,
                       start_date = "2024-11-01", end_date = "2025-02-28")
wx <- fetch_weather_series(sites_win, pars = all_weather_pars,   # omit for the 6 defaults
                           temporal = "daily", workers = 4)
feat <- envirotype_features(wx, windows = 3, stats = c("mean", "sd"),
                            gdd = list(base = 10))
```

Soil is retrieved once per site with `fetch_soilgrids(sites_xy, ...)` (all
SoilGrids properties by default).

## 1.4 Covariate matrix for this (offline) vignette

So the vignette builds without network access, we substitute a simulated
environment-by-covariate matrix in place of `hist_wx$typical` / `feat` above. In
a real run you would use the fetched features directly.

```{r env-covariates}
E <- length(sites)
mk_block <- function(p, sd = 1) {
  m <- matrix(rnorm(E * p, sd = sd), E, p)
  rownames(m) <- sites
  m
}
weather    <- mk_block(6);  colnames(weather)    <- c("mean_temp","tmax","tmin",
                                                       "precip","rad","rh")
soil       <- mk_block(4);  colnames(soil)       <- c("clay","phh2o","soc","cec")
management <- mk_block(2);  colnames(management) <- c("seeding","N_rate")
geography  <- mk_block(3);  colnames(geography)  <- c("lat","lon","elev")
```

## 1.2 Environment kernels

`build_environment_kernels()` turns each covariate block into a Gaussian kernel
at several bandwidths, handling missingness and coverage checks.

```{r env-kernels}
kernels <- build_environment_kernels(
  weather = weather, soil = soil, management = management, geography = geography,
  environments = sites,
  bandwidth_multipliers = c(0.5, 1, 2),
  missing_action = "impute", impute = "median")
names(kernels)
```

## 1.3 Data-driven environmental covariance $\Sigma_E$

If you have **historical MET yield data**, calibrate the kernel mixture against
the empirically observed cross-environment genetic correlations. Interaction
kernels are admitted only on cross-validated / bootstrap **evidence**, never by
default.

```{r calibrate}
# Simulated historical MET: genotype x environment yields over several years.
hist_years <- 2018:2024
historical <- do.call(rbind, lapply(hist_years, function(y) {
  g <- sprintf("G%02d", 1:40)
  expand <- expand.grid(genotype = g, environment = sites,
                        stringsAsFactors = FALSE)
  expand$year  <- y
  expand$value <- rnorm(nrow(expand))
  expand
}))

cal <- calibrate_environment_covariance(
  kernels, historical = historical, year_col = "year",
  genotype_col = "genotype", environment_col = "environment",
  value_col = "value", identity_weight = 0.10, n_boot = 100L, seed = 1)

Sigma_E <- cal$Sigma_E              # E x E environmental covariance
cal$weights                          # fitted kernel weights (on the simplex)
cal$interaction_evidence             # per-interaction CV / bootstrap evidence
```

When no historical target is available, `calibrate_environment_covariance()`
returns the identity as a neutral central covariance and keeps interaction
kernels as *unweighted structural-uncertainty scenarios* — an honest fallback
rather than invented weights.

## 1.4 Enviromic kinship and mega-environments

```{r mega}
# Enviromic similarity D among sites (Gaussian kernel over covariates).
env_cov <- cbind(weather, soil, geography)
D <- build_environment_relationship(env_cov, source = "enviromic",
                                    kernel = "gaussian")

# Cluster sites into mega-environments (silhouette/stability-screened variant:
# infer_mega_environments()).
mega <- cluster_environments(D, n_clusters = 3, Sigma_E = Sigma_E, seed = 1)
mega$membership                       # site -> mega-environment label
mega$clusters                         # sites grouped by mega-environment

# Optionally pick a representative subset of sites under a plot budget.
select_environments(D, n = 5, method = "representative")
```

---

# Module 2 — Genetic relationship matrices

**Goal:** a positive-definite, invertible relationship matrix `G` over the
genotypes, because every reliability/PEV computation needs $G^{-1}$.

## 2.1 Genomic relationship (VanRaden)

```{r grm}
# Simulated 0/1/2 marker dosages for the 300 testcross hybrids.
markers <- matrix(rbinom(n_hybrids * n_markers, 2, 0.3),
                  n_hybrids, n_markers,
                  dimnames = list(hybrids, sprintf("snp%04d", 1:n_markers)))

G <- build_relationship_matrix(markers = markers, type = "genomic",
                               method = "vanraden")
```

## 2.2 Hybrid / testcross GRM with dominance (optional)

For testcrosses, an additive **and** a dominance relationship can matter. Build
a hybrid GRM from parental genotypes and combine additive + dominance:

```{r hybrid, eval = FALSE}
# G_parents: GRM among inbred parents; crosses: female/male parent of each hybrid.
G_A <- build_hybrid_relationship(G_parents = G_parents, crosses = crosses)
# A dominance matrix G_D can be built the same way (Vitezica) and blended:
G  <- combine_relationship_matrices(list(additive = G_A, dominance = G_D),
                                    weights = c(0.8, 0.2))
```

## 2.3 Make it invertible

```{r tune}
G <- tune_relationship_matrix(G, method = "auto")   # blend/bend/rcn to PD
```

---

# Module 3 — Sparse allocation across environments

**Goal:** decide who is tested where. First check feasibility, then let the
package suggest a connected set of **common treatments**, then allocate.

## 3.1 Feasibility and per-site capacity

```{r capacity}
# Sites capped at their physical plot limits (checks included).
capacity <- c(Ecu_Daule = 220, Ecu_Samborondon = 220, Ecu_Babahoyo = 220,
              Ecu_Vinces = 220, Col_Saldana = 200, CIAT_Palmira = 200,
              Per_Chiclayo = 182)

# For an uncapped site, let the package pick the plot number that maximises
# accuracy within an affordable interval:
suggest_site_capacity(G, Sigma_E, candidate_plots = c(150, 180, 200, 220),
                      scope = "single", other_capacity = 200, n_sim = 20,
                      select = "max", seed = 1)$recommended_plots

# Smallest safe common-block size k (buffer for checks):
suggest_safe_k(hybrids, sites, buffer = 3L)
```

## 3.2 Auto-suggested common treatments

`suggest_common_treatments()` chooses **how many** and **which** genotypes to
grow everywhere, balancing (a) seed feasibility across all sites, (b) coverage
of every family and genetic group, and (c) enough overlap to estimate GxE and
cross-site genetic correlations.

```{r commons}
treatment_info <- data.frame(Treatment = hybrids, Family = families,
                             stringsAsFactors = FALSE)
seed_info      <- data.frame(Treatment = hybrids,
                             SeedAvailable = round(runif(n_hybrids, 30, 300)))

cc <- suggest_common_treatments(
  G, treatment_info = treatment_info, seed_info = seed_info,
  seed_required_per_plot = 10,          # grams per plot (Ecuador basis)
  Sigma_E = Sigma_E, n_test_entries = 200, seed = 1)

cc$n_common                             # suggested number
common <- cc$selected                   # the chosen genotypes
cc$rationale                            # why: seed, family, group, connectivity
```

## 3.3 Allocate — M3 (coverage) and M4 (equal replication)

```{r allocate}
# M3: random-balanced, prioritises complete coverage of the TPG.
alloc_m3 <- allocate_sparse_met(
  treatments = hybrids, environments = sites,
  allocation_method = "random_balanced",
  n_test_entries_per_environment = capacity - 20,   # leave room for checks
  common_treatments = common,
  allocation_group_source = "Family", treatment_info = treatment_info,
  Sigma_E = Sigma_E, seed = 1)

alloc_m3$allocation_matrix[1:5, ]       # J x E incidence
alloc_m3$environment_sizes
alloc_m3$summary

# M4: equireplicate, enforces equal network replication when the requested
# degree sequence is Gale-Ryser realisable.
alloc_m4 <- allocate_sparse_met(
  treatments = hybrids, environments = sites,
  allocation_method = "equireplicate",
  n_test_entries_per_environment = rep(180L, E),
  target_replications = 3, common_treatments = common, seed = 1)
```

`A <- alloc_m3$allocation_matrix` is the object every optimisation below
consumes.

```{r A}
A <- alloc_m3$allocation_matrix
```

---

# Module 4 — Seed-aware replication

**Goal:** convert the single network-wide seed inventory into a feasible
replication plan per site, respecting grams-per-plot (which differs by site:
direct seeding vs transplanting).

```{r seed}
seed_df <- data.frame(Treatment = hybrids,
                      SeedAvailable = seed_info$SeedAvailable)

rep_plan <- assign_replication_by_seed(
  treatments = hybrids,
  check_treatments = common,             # commons act as replicated checks
  seed_available = seed_df,
  seed_required_per_plot = 10,
  replication_mode = "p_rep",            # partially replicated
  desired_replications = 2,
  priority = "seed_available")
head(rep_plan)
```

---

# Module 5 — Within-environment field design

**Goal:** lay out each site. Two engines cover the common cases; both return a
`field_book` with `Row`/`Column` (for the field map in Module 9) and feed an
efficiency factor $\epsilon_e \in (0,1]$ back into the coupled optimisation.

## 5.1 Block-based repeated-check design

```{r famoptg}
fb_block <- met_prep_famoptg(
  check_treatments   = common[1:4],
  check_families     = rep("CHK", 4),
  p_rep_treatments   = hybrids[1:60],
  p_rep_reps         = 2,
  p_rep_families     = families[1:60],
  unreplicated_treatments = hybrids[61:180],
  n_rows = 20, n_cols = 12, seed = 1)

eff_block <- met_evaluate_famoptg_efficiency(
  fb_block$field_book, n_rows = 20, n_cols = 12,
  check_treatments = common[1:4])
```

## 5.2 Alpha row-column design

```{r alpha}
fb_alpha <- met_alpha_rc_stream(
  check_treatments = common[1:4], check_families = rep("CHK", 4),
  entry_treatments = hybrids[1:180], entry_families = families[1:180],
  n_reps = 2, n_rows = 20, n_cols = 18, seed = 1)
```

`compare_spatial_models()` can score IID / AR1×AR1 / P-spline error models to
pick the analysis model each layout best supports.

---

# Module 6 — Coupled optimisation & information

This is the scientific core: the **two-level information matrix** that couples
the allocation (Module 3) with the within-site efficiency (Module 5). With
G$\times$E effects $u \sim \mathcal{N}\!\big(0,\ \sigma_g^2\,\Sigma_E \otimes G\big)$
and per-cell information $d_{je} = r_{je}\,\epsilon_e / \sigma_e^2$, absorbing
the environment means gives the effect information matrix

$$
C_{uu} \;=\; D \;-\; D X (X^\top D X)^{-1} X^\top D \;+\;
\sigma_g^{-2}\,\big(\Sigma_E^{-1} \otimes G^{-1}\big),
$$

from which the across-TPE reliability of line $j$ is

$$
\mathrm{CD}_j \;=\; 1 - \frac{\mathrm{PEV}_j}{\sigma_g^2\, G_{jj}\,\bar{\Sigma}_E},
\qquad
\mathrm{PEV}_j \;=\; \frac{1}{E^2}\, a_j^\top C_{uu}^{-1} a_j .
$$

## 6.1 Reliability of the whole design

```{r info}
eff <- setNames(rep(0.9, E), sites)     # e.g. from Module 5 evaluations
info <- met_information(A, G = G, Sigma_E = Sigma_E,
                        sigma_g2 = 1, sigma_e2 = 1, env_efficiency = eff)
info$CDmean                              # mean across-TPE reliability
info$mean_PEV
```

## 6.2 Exact multi-trait index reliability

For a selection index $H = w^\top g$ over several traits with genetic covariance
$\Sigma_T$ and residual covariance $R_T$, `met_information_mt()` solves the
**full multi-trait model** via the Hayes–Hill canonical transform — simultaneously
diagonalising $R_T$ and $\Sigma_T$ ($Q R_T Q^\top = I$, $Q \Sigma_T Q^\top =
\mathrm{diag}(\lambda)$) so the problem decouples into $T$ single-trait solves.
The index reliability is

$$
\mathrm{CD}_H \;=\; 1 - \frac{w^\top \mathrm{PEV}\, w}{\,G_{jj}\,\bar{\Sigma}_E\;
w^\top \Sigma_T\, w\,},
$$

which is exact, not an approximation (see the validation appendix).

```{r mt}
Sigma_T <- matrix(c(1, 0.3, 0.3, 0.6), 2,
                  dimnames = list(c("yield","quality"), c("yield","quality")))
R_T     <- matrix(c(1, 0.2, 0.2, 0.8), 2)
w       <- desired_gain_weights(desired_gain = c(yield = 1, quality = 0.5),
                                gen_cov = Sigma_T, method = "pesek_baker")
info_mt <- met_information_mt(A, G = G, Sigma_E = Sigma_E,
                              Sigma_T = Sigma_T, R_T = R_T, index_weights = w)
info_mt$CDmean_index
```

## 6.3 Objective, optimisation, and robustness

```{r optimize}
# A configurable objective: reliability + expected gain - plot cost.
obj <- design_objective(A, G = G, Sigma_E = Sigma_E,
                        prop = 0.1, cost_per_plot = 1)

# Improve the allocation by simulated-annealing exchanges while preserving the
# per-site margins; aggregate over variance scenarios by conditional value at
# risk (tail-robust).
opt <- optimize_design(
  A, G = G, Sigma_E = Sigma_E, objective = list(prop = 0.1),
  preserve = "margins",
  robust = robust_scenarios(sigma_e2 = c(0.5, 1, 2)),
  robust_aggregate = "cvar", cvar_alpha = 0.25, seed = 1)
opt$allocation_matrix                    # improved incidence
```

## 6.4 Pareto frontier (budget vs. gain)

```{r pareto}
pf <- pareto_designs(A, G = G, Sigma_E = Sigma_E,
                     cost_weights = seq(0, 1, by = 0.25), benefit = "gain")
plot_pareto_frontier(pf, x = "cost")
```

---

# Module 7 — Simulation

**Goal:** validate the analytic reliability with an independent Monte-Carlo
GBLUP: simulate true breeding values, predict from the sparse design, and
measure realised accuracy, selection gain, and top-fraction coincidence.

```{r simulate}
sim <- simulate_met(A, G = G, Sigma_E = Sigma_E,
                    n_sim = 200L, select_fraction = 0.10, seed = 1)
c(accuracy = sim$accuracy_mean, gain = sim$gain_mean,
  coincidence = sim$common_selected_mean, CDmean = sim$CDmean)
```

The Monte-Carlo `accuracy_mean` should track the analytic `CDmean` from Module
6 — a built-in consistency check between the two independent code paths.

---

# Module 8 — Benchmarking

**Goal:** show the sparse design is worth it. Build a panel of reference designs
and compare them on the *same* simulated data (paired, common random numbers),
with confidence intervals, decision stability, and tail risk.

```{r benchmark}
designs <- make_met_benchmark_designs(
  treatments = hybrids, environments = sites,
  n_test_entries_per_environment = rep(180L, E),
  target_replications = 3, common_treatments = common, seed = 1)
# designs is a named list: complete / random / M3 / M4 / (your optimised) ...

bench <- benchmark_met_designs(
  designs, G = G, Sigma_E = Sigma_E,
  sigma_e2 = c(0.5, 1, 2),                 # scenarios
  environment_dropout = list(none = character(), drop_peru = "Per_Chiclayo"),
  cost_per_plot = 1, n_boot = 200L, seed = 1)
bench                                       # per-design accuracy/gain CIs, CVaR, cost

# Stability of repeated common-set selections (Jaccard).
summarize_design_stability(
  common_sets = replicate(5, suggest_common_treatments(
    G, treatment_info, seed_info, seed_required_per_plot = 10,
    n_test_entries = 200)$selected, simplify = FALSE))
```

`benchmark_environment_models()` and `benchmark_environment_missingness()`
extend this to the enviromic side: do the calibrated kernels recover a known
covariance, and how do they degrade under missing weather/soil data?

---

# Module 9 — Field books & field maps

**Goal:** the operational deliverable — a field book per site and one combined
book, plus a field map for planting.

## 9.1 Per-site and combined field books

```{r fieldbook}
# Collect each site's within-environment field book (Module 5). Here we reuse
# one layout for illustration; in practice build one per site.
field_books <- setNames(lapply(sites, function(s) fb_alpha$field_book), sites)

combined <- combine_met_fieldbooks(
  field_books,
  replication_modes = setNames(rep("p_rep", E), sites),
  sparse_method = "random_balanced",
  common_treatments = common)

# Individual site book:
subset(combined, Environment == "Col_Saldana")[1:8, ]
# Combined book (all sites, with Environment / IsCommonTreatment flags):
head(combined)
write.csv(combined, "combined_field_book.csv", row.names = FALSE)
```

## 9.2 Field map

The field book carries `Row` / `Column`, so a map is a tile plot with checks and
common treatments highlighted.

```{r fieldmap}
fb <- subset(combined, Environment == "Col_Saldana")
fb$kind <- ifelse(fb$IsCommonTreatment, "common",
           ifelse(is.na(fb$Treatment), "empty", "entry"))
with(fb, {
  plot(Column, Row, type = "n", asp = 1, main = "Col_Saldana field map",
       xlab = "Column", ylab = "Row")
  cols <- c(entry = "grey85", common = "#0C4C80", empty = "white")
  rect(Column - .5, Row - .5, Column + .5, Row + .5,
       col = cols[kind], border = "grey60")
})
```

---

# One-call pipeline

For routine use, `plan_sparse_met_design()` chains allocation → seed replication
→ within-site design → combined field book with sensible defaults; and
`run_design_strategy()` wraps environment selection + allocation + evaluation.

```{r oneshot}
plan <- plan_sparse_met_design(
  treatments = hybrids, environments = sites,
  allocation_method = "random_balanced",
  n_test_entries_per_environment = capacity - 20,
  common_treatments = common, seed = 1)
names(plan)
```

---

# Appendix — Numerical validation

The two hardest results were verified against independent full mixed-model
solves (reproducible in `oracle_audit.py`):

| Quantity | Independent reference | Max discrepancy |
|---|---|---|
| `met_information` (absorption form) | full augmented $[\mu;\,u]$ MME inverse | **1.1e-16** |
| `met_information_mt` (canonical transform) | direct $J\cdot E\cdot T$ multi-trait MME | **3.3e-16** |
| `.project_simplex` | scipy SLSQP projection | 3.6e-7 |
| `.fit_kernel_simplex` | scipy constrained least squares | 1.9e-9 |
| `.benchmark_psd_root` | reconstruct `ZZ'` (incl. rank-deficient) | 0–9e-16 |
| Gale–Ryser feasibility | bipartite degree-sequence realisability | exact |

In short: the coupled information matrix and its exact multi-trait extension are
correct to floating-point precision — the reliabilities this pipeline reports
are the same numbers a full mixed-model analysis would return.
```

